---
title: "DE"
author: "Zoe Dellaert"
date: "2024-10-05"
output:
  html_document: default
  github_document: default
---

## Differential expression analysis of LCM RNA Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("genefilter")
require("DESeq2")
require("apeglm")
require("ashr")
require("ggplot2")
require("vsn")
require("hexbin")
require("pheatmap")
require("RColorBrewer")
require("EnhancedVolcano")
require("rtracklayer")
require("tidyverse")

sessionInfo() #provides list of loaded packages and version of R.
```

```{r}
#set standard output directory for figures
outdir <- "../output_RNA/differential_expression"

save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg=NULL) {
  print(plot)

  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}

# Specify colors
ann_colors = list(Tissue = c(OralEpi = "palegreen3" ,Aboral = "mediumpurple1"))
```


## Read and clean count matrix and metadata 

Read in raw count data
```{r}
counts_raw <- read.csv("../output_RNA/stringtie-GeneExt/LCM_RNA_gene_count_matrix.csv", row.names = 1) #load in data
```

gene_id,LCM_15,LCM_16,LCM_20,LCM_21,LCM_26,LCM_27,LCM_4,LCM_5,LCM_8,LCM_9

Read in metadata 
```{r}
meta <- read.csv("../data_RNA/LCM_RNA_metadata.csv") %>%
            dplyr::arrange(Sample) %>%
            mutate(across(c(Tissue, Fragment, Section_Date, LCM_Date), factor)) # Set variables as factors 

meta$Tissue <- factor(meta$Tissue, levels = c("OralEpi","Aboral")) #we want OralEpi to be the baseline
```

Read in gff for gene coordinates

```{r}
gff <- import("../references/Pocillopora_acuta_HIv2.genes.gff3")
gff_transcripts <- as.data.frame(gff) %>% filter(type == "transcript") %>%
                                          select(c(seqnames,start,end,width,strand,ID)) %>%
                                          dplyr::rename("chromosome"=seqnames, "query"=ID)
```


Data sanity checks!
```{r}
stopifnot(all(meta$Sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$Sample == colnames(counts_raw))) #are they the same in the same order?
```

## pOverA filtering to reduce dataset

```{r}
ffun<-filterfun(pOverA(0.49,10))  # Keep genes expressed at 10+ counts in at least 50% of samples
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

There are now 14464 genes in the filtered dataset.

Data sanity checks:
```{r}
all(meta$Sample %in% colnames(filtered_counts)) #are all of the sample names in the metadata column names in the gene count matrix?
all(meta$Sample == colnames(filtered_counts))  #are they the same in the same order? Should be TRUE
```

## [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

Create DESeq object and run DESeq2
```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ Fragment + Tissue)

dds <- DESeq(dds)
```

 ### Extract results for Aboral vs. OralEpi contrast

```{r}
res <- results(dds, contrast = c("Tissue","Aboral","OralEpi"))
resLFC <- lfcShrink(dds, coef="Tissue_Aboral_vs_OralEpi", res=res, type = "apeglm")
```

### Extract results for adjusted p-value < 0.05

```{r}
res <- resLFC

resOrdered <- res[order(res$pvalue),]# save differentially expressed genes

DE_05 <- as.data.frame(resOrdered) %>% filter(padj < 0.05 & abs(log2FoldChange) > 1)
DE_05_Up <- DE_05 %>% filter(log2FoldChange > 0) #Higher in Aboral, Lower in OralEpi
DE_05_Down <- DE_05 %>% filter(log2FoldChange < 0) #Lower in Aboral, Higher in OralEpi

nrow(DE_05)
nrow(DE_05_Up) #Higher in Aboral, Lower in OralEpi
nrow(DE_05_Down) #Lower in Aboral, Higher in OralEpi
```

```{r}
write.csv(as.data.frame(resOrdered), 
          file = file.path(outdir, "DESeq_results.csv"))

write.csv(DE_05, 
          file = file.path(outdir, "DEG_05.csv"))
```

### Visualizing Differential Expression

```{r}
EnhancedVolcano(res, 
    lab = ifelse(res$padj < 0.05, rownames(res), ""),
    x = "log2FoldChange", 
    y = "pvalue"
)
```

## Plots
```{r}
plotMA(results(dds, contrast = c("Tissue","Aboral","OralEpi")), ylim=c(-20,20))
plotMA(res, ylim=c(-20,20))
```

### Log2 Fold Change Comparison

```{r}
resultsNames(dds)

# because we are interested in the comparison and not the intercept, we set 'coef=2'
resNorm <- lfcShrink(dds, coef="Tissue_Aboral_vs_OralEpi", type="normal")
resAsh <- lfcShrink(dds, coef="Tissue_Aboral_vs_OralEpi", type="ashr")

par(mfrow=c(1,3), mar=c(4,4,2,1))
xlim <- c(1,1e5); ylim <- c(-20,20)
plotMA(res, xlim=xlim, ylim=ylim, main="apeglm")
plotMA(resNorm, xlim=xlim, ylim=ylim, main="normal")
plotMA(resAsh, xlim=xlim, ylim=ylim, main="ashr")
```

```{r}
plotCounts(dds, gene=which.max(res$log2FoldChange), intgroup="Tissue")
plotCounts(dds, gene=which.min(res$log2FoldChange), intgroup="Tissue")
```

Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)
rld <- rlog(dds, blind=FALSE)
ntd <- normTransform(dds) # this gives log2(n + 1)

meanSdPlot(assay(vsd), main = "vsd")
meanSdPlot(assay(rld))
meanSdPlot(assay(ntd))

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

Will move forward with vst transformation for visualizations

## Heatmap of count matrix

```{r}
heatmap_metadata <- as.data.frame(colData(dds)[,c("Tissue","Fragment")])

#view all genes
pheatmap(assay(vsd), cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)),
         annotation_colors = ann_colors,color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200))

#view highest count genes
select <- order(rowMeans(counts(dds,normalized=TRUE)),
                decreasing=TRUE)[1:20]

pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)),
         annotation_colors = ann_colors, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200))


#view most significantly differentially expressed genes

select <- order(res$padj)[1:20]

pheatmap(assay(vsd)[select,], cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2, annotation_col=(heatmap_metadata%>% select(Tissue)),
         annotation_colors = ann_colors,color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200))
```

## Heatmap of the sample-to-sample distances
```{r}
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$Tissue, vsd$Fragment, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)
```

## Principal component plot of the samples

```{r}
pcaData <- plotPCA(vsd, intgroup=c("Tissue", "Fragment"), returnData=TRUE, ntop = 14464)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot(pcaData, aes(PC1, PC2, color=Tissue, shape=Fragment)) +
  geom_point(size=2) +
  scale_color_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3"))+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

save_ggplot(PCA, "PCA_allgenes")

PCA_small <- ggplot(pcaData, aes(PC1, PC2, color=Tissue)) +
  geom_point(size=2) +
  scale_color_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3"))+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

ggsave(filename = paste0(outdir,"/PCA_allgenes_small", ".png"), plot = PCA_small, width = 4, height = 2.5, units = "in", dpi = 300)
```

```{r}
pcaData <- plotPCA(vsd, intgroup=c("Tissue", "Fragment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot(pcaData, aes(PC1, PC2, color=Tissue, shape=Fragment)) +
  geom_point(size=2) +
  scale_color_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3"))+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

save_ggplot(PCA, "PCA")

PCA_small <- ggplot(pcaData, aes(PC1, PC2, color=Tissue)) +
  geom_point(size=2) +
  scale_color_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3"))+
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

ggsave(filename = paste0(outdir,"/PCA_small", ".png"), plot = PCA_small, width = 4, height = 2.5, units = "in", dpi = 300)
```

Clearly, the majority of the variance in the data is explained by tissue type!

## Annotation data

Download annotation files from genome website
```{bash, eval=FALSE}

# wget files
wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.Conserved_Domain_Search_results.txt.gz

wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.EggNog_results.txt.gz

wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.KEGG_results.txt.gz

# move to references direcotry
mv *gz ../references

# unzip files
gunzip ../references/*gz
```

```{r}
EggNog <- read.delim("../references/Pocillopora_acuta_HIv2.genes.EggNog_results.txt") %>% dplyr::rename("query" = X.query)

CDSearch <- read.delim("../references/Pocillopora_acuta_HIv2.genes.Conserved_Domain_Search_results.txt", quote = "") %>% dplyr::rename("query" = X.Query)

KEGG <- read.delim("../references/Pocillopora_acuta_HIv2.genes.KEGG_results.txt", header = FALSE) %>% dplyr::rename("query" = V1, "KeggTerm" = V2)
```

```{r}
DE_05$query <- rownames(DE_05)
DE_05_annot <- DE_05 %>% left_join(CDSearch) %>% select(query,everything())
DE_05_eggnog <- DE_05 %>% left_join(EggNog) %>% select(query,everything())

write.csv(as.data.frame(DE_05_eggnog), file=paste0(outdir,"/DE_05_eggnog_annotation.csv"))

annot_all <- as.data.frame(rownames(dds)) %>% dplyr::rename("query" = `rownames(dds)`) %>% left_join(CDSearch)

eggnog_all <- as.data.frame(rownames(dds)) %>% dplyr::rename("query" = `rownames(dds)`) %>% left_join(EggNog)
```

```{r}
gene_labels <- eggnog_all %>% select(query,PFAMs) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) %>% #replace NAs with "" for labelling purposes
  separate(PFAMs, into = c("PFAMs", "rest of name"), sep = ",(?=.*?,)", extra = "merge")
  
#view most significantly differentially expressed genes

select <- order(res$padj)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[select,"PFAMs"], fontsize_row = 5)
top50_DE
save_ggplot(top50_DE, "top50_DE")

#view genes Higher in Aboral, Lower in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange,decreasing = TRUE)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_Aboral <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[select,"PFAMs"], fontsize_row = 5)
up_Aboral
save_ggplot(up_Aboral, "up_Aboral")

#view genes Lower in Aboral, Higher in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_OralEpi <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[select,"PFAMs"], fontsize_row = 5)
up_OralEpi
save_ggplot(up_OralEpi, "up_OralEpi")
```

## Genes of Interest

### Single cell marker genes (Levy et al 2021)
```{r}
MarkerGenes <- read.csv("../references/Pacuta_MarkerGenes_Levy2021.csv") %>% dplyr::rename("query" = 1, "List" = 2, "definition" = 3) %>% filter(List !="Toolkit")

MarkerGenes_broc <- read.csv("../output_RNA/marker_genes/Pacuta_Spis_Markers_pairs.csv") %>% select(protein_id_spB,cluster,Standardized_Name_spA ) %>% dplyr::rename("query" = 1, "List" = 2)

MarkerGenes$def_short <- ifelse(nchar(MarkerGenes$definition) > 20, 
                            paste0(substr(MarkerGenes$definition, 1, 17), "..."), 
                            MarkerGenes$definition)
```

### My marker genes (p compressa)

```{r}
Pcomp_MarkerGenes_broc <- read.csv("../output_RNA/marker_genes/combined_Pcomp_markers.csv") %>% select(protein_id_spA,cluster_guess_std,protein_id_spB) %>% dplyr::rename("Pcomp_gene" = protein_id_spA,"query" = protein_id_spB, "List" = cluster_guess_std) 

Pcomp_MarkerGenes_broc <- Pcomp_MarkerGenes_broc %>% filter(grepl("Pocillopora",query)&!grepl("unidentified",List)) %>% distinct() 

Pcomp_MarkerGenes_broc_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% select(query,everything()) %>% left_join(Pcomp_MarkerGenes_broc) 

Pcomp_MarkerGenes_broc_all_res_Pcomp <- Pcomp_MarkerGenes_broc_all_res %>%
  group_by(Pcomp_gene) %>%
  slice_min(order_by = padj, with_ties = FALSE) %>%
  ungroup()

Pcomp_MarkerGenes_broc_all_res_Pacuta <- Pcomp_MarkerGenes_broc_all_res %>% group_by(query,baseMean,log2FoldChange,lfcSE,pvalue,padj) %>% summarize(List = paste(List, collapse = ";"), .groups="drop")

write.csv(Pcomp_MarkerGenes_broc_all_res_Pcomp, "../output_RNA/differential_expression/Pcomp_snRNA_marker_expr_PacutaLCM.csv")
```

```{r}
Pdam_MarkerGenes_broc <- read.csv("../output_RNA/marker_genes/combined_Pdam_markers.csv") %>% select(protein_id_spA,Standardized_Name_spA,protein_id_spB) %>% dplyr::rename("Pdam_gene" = protein_id_spA,"query" = protein_id_spB, "List" = Standardized_Name_spA) 

Pdam_MarkerGenes_broc <- Pdam_MarkerGenes_broc %>% filter(grepl("Pocillopora",query)&!grepl("unidentified",List)) %>% distinct() 

Pdam_MarkerGenes_broc_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% 
  select(query,everything()) %>% left_join(Pdam_MarkerGenes_broc) 

Pdam_MarkerGenes_broc_all_res_Pdam <- Pdam_MarkerGenes_broc_all_res %>%
  group_by(Pdam_gene) %>%
  slice_min(order_by = padj, with_ties = FALSE) %>%
  ungroup()

Pdam_MarkerGenes_broc_all_res_Pacuta <- Pdam_MarkerGenes_broc_all_res %>% group_by(query,baseMean,log2FoldChange,lfcSE,pvalue,padj) %>% summarize(List = paste(List, collapse = ";"), .groups="drop")

write.csv(Pdam_MarkerGenes_broc_all_res_Pdam, "../output_RNA/differential_expression/Pdam_snRNA_marker_expr_PacutaLCM.csv")
```

```{r}
Mcap_MarkerGenes_broc <- read.csv("../output_RNA/marker_genes/combined_Mcap_markers.csv") %>% select(protein_id_spA,cluster_guess_std,protein_id_spB) %>% dplyr::rename("Mcap_gene" = protein_id_spA,"query" = protein_id_spB, "List" = cluster_guess_std) 

Mcap_MarkerGenes_broc <- Mcap_MarkerGenes_broc %>% filter(grepl("Pocillopora",query)&!grepl("unidentified",List)) %>% distinct() 

Mcap_MarkerGenes_broc_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% 
  select(query,everything()) %>% left_join(Mcap_MarkerGenes_broc) 

Mcap_MarkerGenes_broc_all_res_Mcap <- Mcap_MarkerGenes_broc_all_res %>%
  group_by(Mcap_gene) %>%
  slice_min(order_by = padj, with_ties = FALSE) %>%
  ungroup()

Mcap_MarkerGenes_broc_all_res_Pacuta <- Mcap_MarkerGenes_broc_all_res %>% group_by(query,baseMean,log2FoldChange,lfcSE,pvalue,padj) %>% summarize(List = paste(List, collapse = ";"), .groups="drop")

write.csv(Mcap_MarkerGenes_broc_all_res_Mcap, "../output_RNA/differential_expression/Mcap_snRNA_marker_expr_PacutaLCM.csv")
```


### Biomineralization toolkit
```{r}
Biomin <- read.csv("../output_RNA/marker_genes/Pacuta_Biomin_Blast.csv") %>% dplyr::rename("query" = Pocillopora_acuta_best_hit) %>% select(-c(accessionnumber.geneID, Ref))
Biomin_broc <- read.csv("../output_RNA/marker_genes/Pacuta_Biomin_Spis_ortholog.csv") %>% dplyr::rename("query" = Pacuta_gene) %>% select(-c(X,accessionnumber_gene_id, ref))

Biomin <- Biomin %>%
  group_by(query,List) %>%
  summarize(definition = paste(unique(definition), collapse = ","))

Biomin$def_short <- ifelse(nchar(Biomin$definition) > 40, 
                            paste0(substr(Biomin$definition, 1, 37), "..."), 
                            Biomin$definition)

Biomin_filtered_counts <- filtered_counts[(rownames(filtered_counts) %in% Biomin$query),]

Biomin_broc <- Biomin_broc %>%
  group_by(query,List) %>%
  summarize(definition = paste(unique(definition), collapse = ","))

Biomin_broc$def_short <- ifelse(nchar(Biomin_broc$definition) > 40, 
                            paste0(substr(Biomin_broc$definition, 1, 37), "..."), 
                            Biomin_broc$definition)

Biomin_broc_filtered_counts <- filtered_counts[(rownames(filtered_counts) %in% Biomin_broc$query),]

write.csv(Biomin_filtered_counts, "../output_RNA/differential_expression/Biomin_filtered_counts.csv")
```

### Additional gene lists

He_etal_Nvec = He, S., Shao, W., Chen, S. (Cynthia), Wang, T. and Gibson, M. C. (2023). Spatial transcriptomics reveals a cnidarian segment polarity program in Nematostella vectensis. Current Biology 33, 2678-2689.e5.

DuBuc_etal_Nvec = DuBuc, T. Q., Stephenson, T. B., Rock, A. Q. and Martindale, M. Q. (2018). Hox and Wnt pattern the primary body axis of an anthozoan cnidarian before gastrulation. Nat Commun 9, 2007.

- oral/aboral patterning markers in nematostella

```{r}
He_etal_Nvec <- read.csv("../output_RNA/marker_genes/He_etal_nematostella.csv") %>% dplyr::rename("query" = Pacuta_gene) %>% select(-c(X))

He_etal_Nvec$def_short <- gsub("Homeobox protein", "Hox", He_etal_Nvec$Description, ignore.case = TRUE)

DuBuc_etal_Nvec <- read.csv("../output_RNA/marker_genes/Wnt_nematostella.csv") %>% dplyr::rename("query" = Pacuta_gene) %>% select(-c(X))
DuBuc_etal_Nvec$def_short <- DuBuc_etal_Nvec$Gene_Name

HeatStressGenes <- read.csv("../output_RNA/marker_genes/HeatStressGenes_Pacuta.csv") %>% dplyr::rename("query" = Pacuta_gene) %>% select(-c(X))
```


```{r}
DE_05$query <- rownames(DE_05)
resOrdered$query <- rownames(resOrdered)

join_genes_of_interest <- function(df, gene_set) {
  df %>%
    left_join(gene_set, by = "query") %>%
    select(query, everything()) %>%
    drop_na() %>% distinct()
}

DE_05_biomin         <- join_genes_of_interest(DE_05, Biomin)
DE_05_Biomin_broc    <- join_genes_of_interest(DE_05, Biomin_broc)
DE_05_marker         <- join_genes_of_interest(DE_05, MarkerGenes)
DE_05_marker_broc    <- join_genes_of_interest(DE_05, MarkerGenes_broc)
DE_05_He_etal        <- join_genes_of_interest(DE_05, He_etal_Nvec)
DE_05_DuBuc_etal     <- join_genes_of_interest(DE_05, DuBuc_etal_Nvec)

DESeq_biomin         <- join_genes_of_interest(as.data.frame(resOrdered), Biomin)
DESeq_Biomin_broc    <- join_genes_of_interest(as.data.frame(resOrdered), Biomin_broc)
DESeq_marker         <- join_genes_of_interest(as.data.frame(resOrdered), MarkerGenes)
DESeq_marker_broc    <- join_genes_of_interest(as.data.frame(resOrdered), MarkerGenes_broc)
DESeq_He_etal        <- join_genes_of_interest(as.data.frame(resOrdered), He_etal_Nvec)
DESeq_DuBuc_etal     <- join_genes_of_interest(as.data.frame(resOrdered), DuBuc_etal_Nvec)

DE_05_HeatStressGenes <- HeatStressGenes %>% left_join(DE_05, by="query") %>% select(query,everything()) %>% drop_na(baseMean)
DESeq_HeatStressGenes <- HeatStressGenes %>% left_join(as.data.frame(resOrdered), by="query") %>% select(query,everything())  %>% drop_na(baseMean)

write.csv(as.data.frame(DE_05_biomin), file=paste0(outdir,"/DE_05_biomin_annotation.csv"))
write.csv(as.data.frame(DE_05_marker), file=paste0(outdir,"/DE_05_markergene_annotation.csv"))

biomin_all_counts <- as.data.frame(counts(dds)) %>% mutate(query = rownames(dds)) %>% select(query,everything()) %>% left_join(Biomin) 
biomin_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% select(query,everything()) %>% left_join(Biomin) 

Biomin_broc_all_counts <- as.data.frame(counts(dds)) %>% mutate(query = rownames(dds)) %>% select(query,everything()) %>% left_join(Biomin_broc) 
Biomin_broc_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% select(query,everything()) %>% left_join(Biomin_broc) 

markers_all_counts <- as.data.frame(counts(dds)) %>% mutate(query = rownames(dds)) %>% select(query,everything()) %>% left_join(MarkerGenes) 
markers_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% select(query,everything()) %>% left_join(MarkerGenes) 

broc_markers_all_counts <- as.data.frame(counts(dds)) %>% mutate(query = rownames(dds)) %>% select(query,everything()) %>% left_join(MarkerGenes_broc) 
broc_markers_all_res <- as.data.frame(res) %>% mutate(query = rownames(res)) %>% select(query,everything()) %>% left_join(MarkerGenes_broc) 
```

```{r}
#view biomin genes that are differentially expressed

z_scores <- t(scale(t(assay(vsd)[DE_05_biomin$query, ]), center = TRUE, scale = TRUE))
DE_biomin <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_biomin$def_short, fontsize_row = 5)
DE_biomin
save_ggplot(DE_biomin, "DE_biomin")

#view biomin genes that are differentially expressed

z_scores <- t(scale(t(assay(vsd)[DE_05_Biomin_broc$query, ]), center = TRUE, scale = TRUE))
DE_Biomin_broc <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_Biomin_broc$def_short, fontsize_row = 5)
DE_Biomin_broc
save_ggplot(DE_Biomin_broc, "DE_Biomin_broc")

#view marker genes that are differentially expressed

z_scores <- t(scale(t(assay(vsd)[DE_05_marker$query, ]), center = TRUE, scale = TRUE))
DE_marker <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,cutree_rows = 5,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_marker$List, fontsize_row = 4)
DE_marker
save_ggplot(DE_marker, "DE_marker")

DE_05_marker_grouped <- DE_05_marker %>% arrange(List) %>% mutate(List = as.factor(List))

z_scores <- t(scale(t(assay(vsd)[DE_05_marker_grouped$quer, ]), center = TRUE, scale = TRUE))
DE_05_marker_grouped_plot <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_marker_grouped$List, fontsize_row = 5)

DE_05_marker_grouped_plot
save_ggplot(DE_05_marker_grouped_plot, "DE_05_marker_grouped")

#view marker genes that are differentially expressed

z_scores <- t(scale(t(assay(vsd)[DE_05_marker_broc$query, ]), center = TRUE, scale = TRUE))
DE_marker <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,cutree_rows = 5,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_marker_broc$List, fontsize_row = 2)
DE_marker
save_ggplot(DE_marker, "DE_marker_broc")

DE_05_marker_broc_grouped <- DE_05_marker_broc %>% arrange(List) %>% mutate(List = as.factor(List))

z_scores <- t(scale(t(assay(vsd)[DE_05_marker_broc_grouped$quer, ]), center = TRUE, scale = TRUE))
DE_05_marker_broc_grouped_plot <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = DE_05_marker_broc_grouped$List, fontsize_row = 3)

DE_05_marker_broc_grouped_plot
save_ggplot(DE_05_marker_broc_grouped_plot, "DE_05_marker_broc_grouped")
```

### Visualizing Differential Expression

```{r}
Biomin_volcano <- EnhancedVolcano(biomin_all_res, 
    lab = biomin_all_res$def_short,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.01,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 40)

save_ggplot(Biomin_volcano, "Biomin_volcano")

Biomin_broc_volcano <- EnhancedVolcano(Biomin_broc_all_res, 
    lab = Biomin_broc_all_res$def_short,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.01,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 40)

save_ggplot(Biomin_broc_volcano, "Biomin_broc_volcano")

Marker_volcano <- EnhancedVolcano(markers_all_res, 
    lab = markers_all_res$List,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.01,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 60)

save_ggplot(Marker_volcano, "Marker_volcano")

Marker_volcano_names <- EnhancedVolcano(markers_all_res, 
    lab = markers_all_res$def_short,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.01,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 60)

save_ggplot(Marker_volcano_names, "Marker_volcano_names")

Marker_volcano <- EnhancedVolcano(broc_markers_all_res, 
    lab = broc_markers_all_res$List,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.01,
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 60)

save_ggplot(Marker_volcano, "Marker_volcano_broc")

EnhancedVolcano(res, 
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue')

volcano_plain <- EnhancedVolcano(res, 
    lab = NA,
    x = 'log2FoldChange',
    y = 'padj',
    pCutoff = 0.05,
    title="",
    subtitle="",
    drawConnectors = TRUE,
    widthConnectors = 0.75,
    pointSize = 1,
    labSize = 2,boxedLabels = TRUE,max.overlaps = 60)

save_ggplot(volcano_plain, "volcano_plain",width = 4, height = 6, units = "in", dpi = 300)
```

## Trying to annotate un-annotated DE genes:

```{bash, eval = FALSE}
# wget protein sequence reference file
wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.pep.faa.gz

# move to references direcotry
mv *gz ../references

# unzip files
gunzip ../references/*gz

#get the names of all the DEGs from the first column of the DEG csv file
tail -n +2 ../output_RNA/differential_expression/DEG_05.csv | cut -d',' -f1 | tr -d '"' > ../output_RNA/differential_expression/DEG_05_names.csv

#grep this file against the protein fasta file, first with wc -l to make sure the number of lines is correct (should be your number of DEGs)
grep -f ../output_RNA/differential_expression/DEG_05_names.csv ../references/Pocillopora_acuta_HIv2.genes.pep.faa | wc -l

#grep each header with the protein sequence after ("-A 1") and save to a new file
grep  -A 1 -f ../output_RNA/differential_expression/DEG_05_names.csv ../references/Pocillopora_acuta_HIv2.genes.pep.faa > ../output_RNA/differential_expression/DEG_05_seqs.txt
```

### nr database BLAST of DE_05 genes only

On andromeda:

Blastp-ing only the DE genes against the entire nr database (will take a while)

```{bash eval=FALSE}
cd ../scripts
nano DEG_05_blast.sh
```

```{bash eval=FALSE}
#!/bin/bash
#SBATCH --job-name="DE_blast"
#SBATCH -t 240:00:00
#SBATCH --export=NONE
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH --mail-user=zdellaert@uri.edu #your email to send notifications
#SBATCH --mem=500GB
#SBATCH --error=../scripts/outs_errs/"%x_error.%j" #write out slurm error reports
#SBATCH --output=../scripts/outs_errs/"%x_output.%j" #write out any program outpus
#SBATCH --account=putnamlab
#SBATCH --nodes=2 --ntasks-per-node=24

module load BLAST+/2.15.0-gompi-2023a

cd ../output_RNA/differential_expression #set working directory
mkdir blast
cd blast

#nr database location andromeda: /data/shared/ncbi-db/.ncbirc 
# points to current location: cat /data/shared/ncbi-db/.ncbirc
# [BLAST]
# BLASTDB=/data/shared/ncbi-db/2024-11-10

blastp -query ../DEG_05_seqs.txt -db nr -out DEG_05_blast_results.txt -outfmt 0 -evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 10

blastp -query ../DEG_05_seqs.txt -db nr -out DEG_05_blast_results_tab.txt -outfmt 6 -evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 1 \
-max_hsps 1

echo "Blast complete!" $(date)
```

```{bash, eval=FALSE}
sbatch DEG_05_blast.sh
```

```{bash, eval=FALSE}
cd ../output_RNA/differential_expression/blast

wc -l DEG_05_blast_results_tab.txt #3537 of the 3606 genes were annotated

#get just the NCBI database accession numbers for the blast results
cut -f2 DEG_05_blast_results_tab.txt > DEG_05_blast_accessions.txt

#remove any duplicates
sort -u  DEG_05_blast_accessions.txt > unique_DEG_05_blast_accessions.txt

wc -l unique_DEG_05_blast_accessions.txt #3404 of the 3537 annotations were unique

while read acc; do
  curl -s "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=protein&id=$acc&rettype=gp&retmode=text" \
  | grep "DEFINITION" | sed 's/DEFINITION  //g' | awk -v id="$acc" '{print id "\t" $0}'
done < unique_DEG_05_blast_accessions.txt > DEG_05_blast_names.txt

wc -l DEG_05_blast_names.txt #3396 ; unsure why 8 are missing.

join -1 2 -2 1 -t $'\t' <(sort -k2 DEG_05_blast_results_tab.txt) <(sort DEG_05_blast_names.txt) > annotated_DEG_05_blast_results_tab.txt
```

### SwissProt of ALL genes

On unity:

swissprot based on https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/20-Apul-gene-annotation.Rmd and
https://github.com/urol-e5/deep-dive/blob/main/F-Pmea/code/20-Pmea-gene-annotation.Rmd

Steven's notebook post here: https://sr320.github.io/tumbling-oysters/posts/sr320-27-go/

```{bash, eval=FALSE}
mkdir ../references/blast_dbs
```

```{bash, eval=FALSE}
cd ../references/blast_dbs
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
mv uniprot_sprot.fasta.gz uniprot_sprot_r2024_10_02.fasta.gz
gunzip -k uniprot_sprot_r2024_10_02.fasta.gz
rm uniprot_sprot_r2024_10_02.fasta.gz

head uniprot_sprot_r2024_10_02.fasta
echo "Number of Sequences"
grep -c ">" uniprot_sprot_r2024_10_02.fasta
# 572214 sequences
```

```{bash, eval=FALSE}
module load blast-plus/2.14.1

makeblastdb \
-in ../references/blast_dbs/uniprot_sprot_r2024_10_02.fasta \
-dbtype prot \
-out ../references/blast_dbs/uniprot_sprot_r2024_10_02
```

```{bash, eval=FALSE}
cd ../scripts
nano blastp_SwissProt.sh
```

```{bash eval=FALSE}
#!/bin/bash
#SBATCH -t 18:00:00
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=48
#SBATCH --mem=500GB
#SBATCH --export=NONE
#SBATCH --error=../scripts/outs_errs/"%x_error.%j" #write out slurm error reports
#SBATCH --output=../scripts/outs_errs/"%x_output.%j" #write out any program outpus
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH -D /project/pi_hputnam_uri_edu/zdellaert/LaserCoral #set working directory

module load blast-plus/2.14.1

cd references/
mkdir annotation

fasta="Pocillopora_acuta_HIv2.genes.pep.faa"

blastp \
-query $fasta \
-db blast_dbs/uniprot_sprot_r2024_10_02 \
-out annotation/blastp_SwissProt_out.tab \
-evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 1 \
-max_hsps 1 \
-outfmt 6

echo "Blast complete!" $(date)
```

```{bash, eval=FALSE}
cd references/annotation/

tr '|' '\t' <  blastp_SwissProt_out.tab >  blastp_SwissProt_out_sep.tab
```

```{bash, eval=FALSE}
cd ../references/annotation/

curl -H "Accept: text/plain; format=tsv" "https://rest.uniprot.org/uniprotkb/stream?fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_p%2Cgo%2Cgo_id%2Cgo_c%2Cgo_f&format=tsv&query=%28reviewed%3Atrue%29" -o SwissProt-Annot-GO_111524.tsv

wc -l SwissProt-Annot-GO_111524.tsv
#572215
```

All code below based on https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/20-Apul-gene-annotation.Rmd and
https://github.com/urol-e5/deep-dive/blob/main/F-Pmea/code/20-Pmea-gene-annotation.Rmd

Steven's notebook post here: https://sr320.github.io/tumbling-oysters/posts/sr320-27-go/

```{r}
bltabl <- read.csv("../references/annotation/blastp_SwissProt_out_sep.tab", sep = '\t', header = FALSE)

spgo <- read.csv("../references/annotation/SwissProt-Annot-GO_111524.tsv", sep = '\t', header = TRUE)
```

```{r}
annot_tab <- left_join(bltabl, spgo, by = c("V3" = "Entry")) %>%
  select(
    query = V1,
    blast_hit = V3,
    evalue = V13,
    ProteinNames = Protein.names,
    BiologicalProcess = Gene.Ontology..biological.process.,
    GeneOntologyIDs = Gene.Ontology.IDs,
    CellularComponent = Gene.Ontology..cellular.component.,
    MolecularFunction = Gene.Ontology..molecular.function.,
  )

head(annot_tab)
```

```{r}
write.table(annot_tab, 
            file = "../references/annotation/protein-GO.tsv", 
            sep = "\t", 
            row.names = FALSE, 
            quote = FALSE)

DESeq_SwissProt <- as.data.frame(resOrdered) %>% left_join(annot_tab) %>% select(query,everything()) 
DE_05_SwissProt <- DE_05 %>% left_join(annot_tab) %>% select(query,everything()) 

write.csv(as.data.frame(DE_05_SwissProt), file=paste0(outdir,"/DE_05_SwissProt_annotation.csv"))
```

```{r}
DE_05_SwissProt$short_name <- ifelse(nchar(DE_05_SwissProt$ProteinNames) > 30, 
                            paste0(substr(DE_05_SwissProt$ProteinNames, 1, 27), "..."), 
                            DE_05_SwissProt$ProteinNames)

gene_labels <- DE_05_SwissProt %>% 
  select(query,short_name) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

#view most significantly differentially expressed genes

select <- order(res$padj)[1:50]
rownames(res)[select]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 6)
top50_DE
save_ggplot(top50_DE, "top50_DE_SwissProt")

#view most significantly differentially expressed genes by LFC

select <- order(abs(res$log2FoldChange),decreasing = TRUE)[1:50]
rownames(res)[select]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 6)
top50_DE
save_ggplot(top50_DE, "top50_LFC_DE_SwissProt")


#view genes Higher in Aboral, Lower in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange,decreasing = TRUE)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_Aboral <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 5)
up_Aboral
save_ggplot(up_Aboral, "up_Aboral_SwissProt")

#view genes Lower in Aboral, Higher in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_OralEpi <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 5)
up_OralEpi
save_ggplot(up_OralEpi, "up_OralEpi_SwissProt")
```

```{r}
DE_05_SwissProt$short_GO <- ifelse(nchar(DE_05_SwissProt$BiologicalProcess) > 30, 
                            paste0(substr(DE_05_SwissProt$BiologicalProcess, 1, 27), "..."), 
                            DE_05_SwissProt$BiologicalProcess)

gene_labels <- DE_05_SwissProt %>% select(query,short_GO) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

#view most significantly differentially expressed genes

select <- order(res$padj)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 5)
top50_DE
save_ggplot(top50_DE, "top50_DE_Blast2GO")

#view genes Higher in Aboral, Lower in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange,decreasing = TRUE)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_Aboral <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 5)
up_Aboral
save_ggplot(up_Aboral, "up_Aboral_Blast2GO")

#view genes Lower in Aboral, Higher in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_OralEpi <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 5)
up_OralEpi
save_ggplot(up_OralEpi, "up_OralEpi_Blast2GO")
```



### Combined annotations manually

```{r}
Manual <- read.csv(paste0(outdir,"/DE_05_Manual_annotation.csv")) %>% arrange(X) 
Manual <- Manual %>% filter(abs(log2FoldChange) > 1)

Manual$Heatmap_Label <- gsub("Homeobox protein", "Hox", Manual$Heatmap_Label, ignore.case = TRUE)
Manual$Heatmap_Label <- gsub("Protein Wnt", "Wnt", Manual$Heatmap_Label, ignore.case = TRUE)

gene_labels <- Manual %>% 
  select(query,Heatmap_Label) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes
```

```{r}
#view most significantly differentially expressed genes

select <- order(res$padj)[1:50]
rownames(res)[select]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 6)
top50_DE
save_ggplot(top50_DE, "top50_DE_Manual")

#view most significantly differentially expressed genes by LFC

select <- order(abs(res$log2FoldChange),decreasing = TRUE)[1:50]
rownames(res)[select]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 6)
top50_DE
save_ggplot(top50_DE, "top50_LFC_DE_Manual")


#view genes Higher in Aboral, Lower in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange,decreasing = TRUE)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_Aboral <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), labels_col = NA, annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 8.5)
up_Aboral
save_ggplot(up_Aboral, "up_Aboral_Manual")

#view genes Lower in Aboral, Higher in OralEpi, ordered by log2FoldChange
select <- order(res$log2FoldChange)[1:50]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
up_OralEpi <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), labels_col = NA, annotation_colors = ann_colors,
         labels_row =gene_labels[match(rownames(res)[select],(gene_labels$query)),2], fontsize_row = 8.5)
up_OralEpi
save_ggplot(up_OralEpi, "up_OralEpi_Manual")
```

## Expression of certain genes of interest: blast against P. acuta genome

### yin yang Transctiption factor

```{bash, eval=FALSE}
cd ../references

#download the genome protein fasta if you have not already
wget http://cyanophora.rutgers.edu/Pocillopora_acuta/Pocillopora_acuta_HIv2.genes.pep.faa.gz

#unzip file
gunzip Pocillopora_acuta_HIv2.genes.pep.faa.gz
```

In unity

```{bash, eval=FALSE}
salloc -p cpu -c 8 --mem 32G

module load uri/main
module load BLAST+/2.15.0-gompi-2023a

#make blast_dbs directory if you haven't done so above
mkdir blast_dbs
cd blast_dbs

makeblastdb -in ../Pocillopora_acuta_HIv2.genes.pep.faa -out Pacuta_prot -dbtype prot

cd ../../output_RNA/differential_expression/

#make blast output directory if you haven't done so above
mkdir blast
cd blast

nano YinYang.txt
```

add accession numbers of interest:

- nematostella vectensis transcriptional repressor protein YY1 isoforms
  - XP_048585772.1
  - XP_048585773.1
  - XP_048585774.1

- human transcription factor YY2
  - NP_996806.2

- human transcription factor ZFP41 (REX-1)
  - NP_777560.2

```{bash, eval = FALSE}
XP_048585772.1
XP_048585773.1
XP_048585774.1
NP_996806.2
NP_777560.2
```

```{bash, eval = FALSE}
# Read the input file line by line and fetch FASTA sequences
while read -r accession; do
  if [[ -n "$accession" ]]; then
    echo "Fetching $accession..."
    curl -s "https://eutils.ncbi.nlm.nih.gov/entrez/eutils/efetch.fcgi?db=protein&id=${accession}&rettype=fasta&retmode=text" >> "YinYang.fasta"
    echo >> "YinYang.fasta"  # Add a newline between sequences
    sleep 1  # Avoid hitting rate limits
  fi
done < "YinYang.txt"

# run blast with human-readable output
blastp -query YinYang.fasta -db ../../../references/blast_dbs/Pacuta_prot -out YinYang_blast_results.txt -outfmt 0

#looks like there are a lot of matches for each gene! I am going to do a tab search with a very low e-value cutoff:

# run blast with tabular output
blastp -query YinYang.fasta -db ../../../references/blast_dbs/Pacuta_prot -out YinYang_blast_results_tab.txt -outfmt 6 -evalue 1e-25
```

Great! Interestingly, the same gene, Pocillopora_acuta_HIv2___RNAseq.g25242.t1 is the top match for all 5 proteins I searched. Pocillopora_acuta_HIv2___TS.g24434.t1 is the second best match for all 5. That is interesting! Pocillopora_acuta_HIv2___RNAseq.g7583.t1 is also worth looking at. And Pocillopora_acuta_HIv2___TS.g21338.t1 matched all three YY1 isoforms.

```{r}
YinYangs <- c("Pocillopora_acuta_HIv2___RNAseq.g25242.t1",
              "Pocillopora_acuta_HIv2___TS.g24434.t1",
              "Pocillopora_acuta_HIv2___RNAseq.g7583.t1",
              "Pocillopora_acuta_HIv2___TS.g21338.t1")

for (i in 1:length(YinYangs)){
  plotCounts(dds, gene=YinYangs[i], intgroup=c("Tissue"),)
}

as.data.frame(resOrdered)[YinYangs,]
```

Okay! None of these genes are differentially expressed between the tissues. That is interesting and good to know. Pocillopora_acuta_HIv2___RNAseq.g25242.t1 has the highest basal expression of all the potential isoforms of this transcription factor.


### TRP Channels

```{r}
DE_05_SwissProt$short_name <- ifelse(nchar(DE_05_SwissProt$ProteinNames) > 80, 
                            paste0(substr(DE_05_SwissProt$ProteinNames, 1, 77), "..."), 
                            DE_05_SwissProt$ProteinNames)

TRP <- Manual %>% filter(grepl("transient", ProteinNames, ignore.case = TRUE)) 
select <- TRP$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 12)
top50_DE
save_ggplot(top50_DE, "TRP_DE_SwissProt", width = 6.43, height = 4.25, units = "in", dpi = 300)

annot_tab$short_name <- ifelse(nchar(annot_tab$ProteinNames) > 80, 
                            paste0(substr(annot_tab$ProteinNames, 1, 77), "..."), 
                            annot_tab$ProteinNames)

swissprot_labels <- annot_tab  %>% 
  select(query,short_name) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

TRP <- annot_tab %>% filter(grepl("transient", ProteinNames, ignore.case = TRUE))
select1 <- TRP$query
select <- match(select1,rownames(vsd))
select <- select[!is.na(select)]

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = swissprot_labels[match(select1,swissprot_labels$query),2], fontsize_row = 6)
top50_DE
save_ggplot(top50_DE, "TRP_SwissProt")
```

### biomin de heatmap nice

```{r}
DE_05_biomin_filtered <- DE_05_biomin %>% left_join(Manual,by="query" ) #%>% filter(!(Heatmap_Label %in% c("Myosin-9", "Actin, cytoplasmic")))
  
select <- DE_05_biomin_filtered$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_biomin <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
DE_biomin
save_ggplot(DE_biomin, "DE_biomin")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k as needed

# Create a dataframe to manage clusters and reordering
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate heatmap with reordered rows and labels
DE_biomin <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  show_rownames = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8
)
save_ggplot(DE_biomin, "clusters_clean/DE_biomin")

#############################

DE_05_Biomin_broc_filtered <- DE_05_Biomin_broc %>% left_join(Manual,by="query" ) %>% filter(!(Heatmap_Label %in% c("Myosin-9", "Actin, cytoplasmic")))
  
select <- DE_05_Biomin_broc_filtered$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_Biomin_broc <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
DE_Biomin_broc
save_ggplot(DE_Biomin_broc, "DE_Biomin_broc")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k as needed

# Create a dataframe to manage clusters and reordering
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate heatmap with reordered rows and labels
DE_Biomin_broc <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  show_rownames = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8
)
save_ggplot(DE_Biomin_broc, "clusters_clean/DE_Biomin_broc")
```

```{r}
vst_counts <- assay(vsd)["Pocillopora_acuta_HIv2___RNAseq.g15280.t1", ]
df <- as.data.frame(colData(dds))
df$expression <- vst_counts

library(ggplot2)
ggplot(df, aes(x=Tissue, y=expression, group=Fragment, color=Fragment)) +
  geom_point(size=3) +
  geom_line() +
  theme_minimal() +
  labs(title="VST Expression by Tissue", y="VST normalized expression")
```

### Marker DE nice

```{r}
gene_labels <- DE_05_SwissProt %>% 
  select(query,short_name) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

#view marker genes that are differentially expressed
select <- DE_05_marker$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_marker <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), 
                      cluster_rows=TRUE, show_rownames=TRUE,
                      cluster_cols=TRUE, cutree_cols = 2, cutree_rows = 5,
                      annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
                      labels_row = DE_05_marker$List, fontsize_row = 7)
DE_marker
save_ggplot(DE_marker, "DE_marker_2")

#view marker genes that are differentially expressed
select <- DE_05_marker$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_marker <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), 
                      cluster_rows=TRUE, show_rownames=TRUE,
                      cluster_cols=TRUE, cutree_cols = 2, cutree_rows = 5,
                      annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
                      labels_row = paste0(DE_05_marker$List,": ", gene_labels[match(select,(gene_labels$query)),2]), fontsize_row = 7)
DE_marker
save_ggplot(DE_marker, "DE_marker_3")

#view marker genes that are differentially expressed
select <- DE_05_marker_broc$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_marker_broc <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), 
                      cluster_rows=TRUE, show_rownames=TRUE,
                      cluster_cols=TRUE, cutree_cols = 2, cutree_rows = 5,
                      annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
                      labels_row = DE_05_marker_broc$List, fontsize_row = 7)
DE_marker_broc
save_ggplot(DE_marker_broc, "DE_marker_broc_2",height=9)

#view marker genes that are differentially expressed
select <- DE_05_marker_broc$query
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
DE_marker_broc <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), 
                      cluster_rows=TRUE, show_rownames=TRUE,
                      cluster_cols=TRUE, cutree_cols = 2, cutree_rows = 5,
                      annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
                      labels_row = paste0(DE_05_marker_broc$List,": ", gene_labels[match(select,(gene_labels$query)),2]), fontsize_row = 6)
DE_marker_broc
save_ggplot(DE_marker_broc, "DE_marker_broc_3",height=9)
```


### bacteria genes

```{r}
gene_labels <- Manual %>% 
  select(query,Heatmap_Label) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

mucin <- Manual %>% filter(grepl("mucin", ProteinNames, ignore.case = TRUE)|
                             grepl("lectin", ProteinNames, ignore.case = TRUE)|
                             grepl("toll-", ProteinNames, ignore.case = TRUE)|
                             grepl("ZP", ProteinNames, ignore.case = TRUE)|
                             
                             grepl("nitric", ProteinNames, ignore.case = TRUE)|
                             grepl("sulfotransferase", ProteinNames, ignore.case = TRUE)|
                             grepl("Complement ", ProteinNames, ignore.case = TRUE)|
                             grepl("complement activation", BiologicalProcess, ignore.case = TRUE)|
                             grepl("3'-phosphoadenosine 5'-phosphosulfate", BiologicalProcess, ignore.case = TRUE)) %>% 
  filter(Heatmap_Label !="Cnidocyte marker protein (Collectin-11)")

select <- unique(mucin$query)
 
z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "bacteria_DE_SwissProt",height=12)


# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate the heatmap with reordered rows and labels
mucins <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  
  gaps_row = c(43) 
)

# Save the heatmap
save_ggplot(mucins, "clusters_clean/bacteria_DE_SwissProt",height=12)
```



### Symbiont related transporters

```{r}
transport <- Manual %>% filter(grepl("amino acid transport", BiologicalProcess, ignore.case = TRUE)|
                                 grepl("ammonium transmembrane transport", BiologicalProcess, ignore.case = TRUE)|
                                 grepl("carbohydrate transport", BiologicalProcess, ignore.case = TRUE)|
                                 grepl("glycerol transmembrane transport", BiologicalProcess, ignore.case = TRUE)|
                                 grepl("cholesterol transport", BiologicalProcess, ignore.case = TRUE)) %>% 
                                 filter(!grepl("neurotransmitter", BiologicalProcess, ignore.case = TRUE)) %>% 
                                 filter(!grepl("synaptic", BiologicalProcess, ignore.case = TRUE)) %>% 
                                 filter(!grepl("pigmentation", BiologicalProcess, ignore.case = TRUE))# %>% 
                                 #filter(!grepl("Solute carrier", ProteinNames, ignore.case = TRUE))

select <- transport$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "transport_DE_SwissProt")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate the heatmap with reordered rows and labels
transporters_heat <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white", 
  gaps_row = c(10) 
)

# Save the heatmap
save_ggplot(transporters_heat, "clusters_clean/transport_DE_SwissProt")
```

### Solute Carrier (SLC) Family genes

```{r}
solute_all <- DESeq_SwissProt %>% filter(grepl("Solute carrier", ProteinNames, ignore.case = TRUE))
solute_all <- solute_all %>%
  mutate(SLC_family = str_extract(ProteinNames, "Solute carrier family \\d+")) %>%
  mutate(SLC_family = str_replace(SLC_family,"Solute carrier family ", "SLC")) 

nrow(solute_all)

solute_de <- Manual %>% filter(grepl("Solute carrier", ProteinNames, ignore.case = TRUE))
solute_de <- solute_de %>%
  mutate(SLC_family = str_extract(ProteinNames, "Solute carrier family \\d+")) %>%
  mutate(SLC_family = str_replace(SLC_family,"Solute carrier family ", "SLC")) 


solute_de_upAboral <- solute_de %>% filter(log2FoldChange > 1)
nrow(solute_de_upAboral)

solute_de_upOral <- solute_de %>% filter(log2FoldChange < -1)
nrow(solute_de_upOral)

length(unique(solute_all$SLC_family))

length(unique(solute_de_upAboral$SLC_family))
length(unique(solute_de_upOral$SLC_family))
table(solute_de$SLC_family)
```
```{r}
#Basing my annotations on: https://re-solute.eu/knowledgebase/slcfamily and https://slc.bioparadigms.org/

slc_map <- data.frame(
  SLC_family = c(
    "SLC1",
    "SLC12",
    "SLC15",
    "SLC16",
    "SLC22",
    "SLC23",
    "SLC24",
    "SLC25",
    "SLC30",
    "SLC32",
    "SLC35",
    "SLC46",
    "SLC6",
    "SLC7"
  ),
  Function = c(
    "High-affinity glutamate and neutral amino acid transporter family",
    "Electroneutral cation-coupled Cl cotransporter family",
    "Proton oligopeptide cotransporter family",
    "Monocarboxylate transporter family",
    "Organic cation/anion/zwitterion transporter family",
    "Sodium-dependent ascorbic acid transporter family",
    "Na+/(Ca2+-K+) exchanger family",
    "Mitochondrial carrier family",
    "Zinc transporter family",
    "Vesicular inhibitory amino acid transporter family",
    "Nucleoside-sugar transporter family",
    "Folate transporter family",
    "Sodium- and chloride-dependent neurotransmitter transporter family",
    "Cationic amino acid transporter/glycoprotein-associated family"
  ),
  Role = c(
    "Neurotransmission, Glu metabolism, development, aa homeostasis",
    "cell volume, chloride concentration, blood pressure regulation",
    "Immunity, homeostasis of neuropeptides, peptide absoption",
    "glucose homeostasis / metabolism, lipid metabolism",
    "Drug uptake, endocytosis",
    "Absortion and distribution of ascorbic acid",
        "Skin/hair/eye pigmentation, vision, olfaction",
    "Apoptosis, Glucose/Lipid/Amino acid/Urea, metab., ATP synthesis",
    "Zinc homestasis, development, body weight, glucose metabolism",
    "Development, neurotransmission",
    "Virus infection, development, autophagy, cell proliferation",
    "Folate metabolism, Immunity, Drug uptake",
    "Cell differentiation, neurotransmission",
    "Nitric oxide synthesis, redox balance, aa homeostasis"
  ),
  Substrate_Specific = c(
    "Glutamate as Neurotransmitter, anionic, neutral amino acids",
    "Chloride, sodium, potassium",
    "Dipeptides, oligopeptides",
    "Ketone bodies, Lactate, Pyruvate, creatine, aromatic amino acids, T2,T3,T4",
    "Drugs, neurotransmitters, purine and pyrimidine nucleosides, niacin, steroid, carnitine, carnitine derivatives, organic anions",
    "Ascorbic acid",
    "Calcium, Potassium, Sodium",
    "Amino acids, Di- tricarboxylic ions, Co-A, Carnitine/acyl-carnitines, H+, metals, nucleotides",
    "Zinc, Managnese, Lead",
    "GABA, Glycine",
    "Purines, pyrimidines, Thiamine, nuclotide-sugars",
    "Folic acid, Heme, Methotrexate",
    "Neurotransmitters, basic, neutral amino acids",
    "Amino acids"
  ),
  Substrate_Broad = c(
    "amino acids",
    "inorganic ions",
    "peptides",
    "energy metabolites",
    "organic ions",
    "ascorbic acid",
    "inorganic ions",
    "mitochondrial",
    "transition element cations",
    "amino acids; neurotransmitter",
    "nucleosides, nucleotides and nucleotide-sugars",
    "organic ions",
    "neurotransmitter",
    "amino acids"
  )
)
```

```{r}
solute_de_annot <- solute_de %>%
  left_join(slc_map, by = "SLC_family") %>%
  mutate(Direction = ifelse(log2FoldChange > 1, "upAboral",
                            ifelse(log2FoldChange < -1, "upOral", "noChange")))

# Summarize counts by family and direction
plot_data <- solute_de_annot %>%
  filter(Direction != "noChange") %>%
  group_by(SLC_family, Substrate_Broad, Direction) %>%
  summarise(n = n(), .groups = "drop")

ggplot(plot_data, aes(x = SLC_family, y = n, fill = Direction)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Substrate_Broad, scales = "free_x") +
  scale_fill_manual(values = c("upOral" = "darkgreen", "upAboral" = "purple")) +
  theme_minimal() +
  labs(title = "Differentially Expressed SLC Genes by Family",
       x = "SLC Family", y = "Number of DE Genes") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r}
select <- solute_de$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "SLC_DE_SwissProt")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate the heatmap with reordered rows and labels
SLC <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white", 
  gaps_row = c(21) 
)

# Save the heatmap
save_ggplot(SLC, "clusters_clean/SLC_DE_SwissProt")
```

### temp and light sensing

```{r}
sensors <- Manual %>% filter(grepl("cellular response to light stimulus", BiologicalProcess, ignore.case = TRUE)|
                               grepl("detection of mechanical stimulus", BiologicalProcess, ignore.case = TRUE)|
                               grepl("thermoception [GO:0050955]", BiologicalProcess, ignore.case = TRUE)) %>% 
                               filter(!grepl("Solute carrier", ProteinNames, ignore.case = TRUE))

gene_labels <- Manual %>% 
  select(query,Heatmap_Label) %>%
  mutate_all(~ ifelse(is.na(.), "", .)) #replace NAs with "" for labelling purposes

select <- sensors$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "sensors_DE_SwissProt")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  mutate(cluster = factor(cluster, levels = c(1, 3, 2))) %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Generate the heatmap with reordered rows and labels
transporters_heat <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  
  gaps_row = c(11)  # Add breaks only between clusters
)

# Save the heatmap
save_ggplot(transporters_heat, "clusters_clean/sensors_DE_SwissProt")

```

#### stereocilium

```{r}
stereo <- DE_05_SwissProt %>% filter(grepl("stereocilium", CellularComponent, ignore.case = TRUE))

stereo
```

### Zona pellucida

```{r}
Amil_ZP <- DESeq_SwissProt %>% filter(blast_hit=="G8HTB6") %>% pull(query) %>% unique()

PFAM_ZP_domain <- annot_all   %>% filter(grepl("Zona_pellucida",Short.name,ignore.case=TRUE))  %>% 
  #filter(!(query %in% Amil_ZP)) %>% 
  pull(query) %>% unique()

```


### Heat Stress Genes

#### differentially expressed

```{r}
select <- DE_05_HeatStressGenes %>% arrange(response_type) %>% select(query,gene_name,response_type) %>% unique()
rownames(select) <- select$query

z_scores <- t(scale(t(assay(vsd)[select$query, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_row = (select %>% select(response_type)), annotation_colors = ann_colors,
         labels_row = select$gene_name, fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "HeatStress_DE_SwissProt")


# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select$query,
  Heatmap_Label = select$gene_name,
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  annotation_row = (select %>% select(response_type)),
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  
  gaps_row = 4  # Add breaks only between clusters
)


# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/HeatStress_DE_SwissProt")
```


#### all
```{r}
select <- DESeq_HeatStressGenes %>% arrange(response_type) %>% select(query,gene_name,response_type) %>%
    group_by(query) %>% #collapse duplicate rows
  summarise(
    gene_name = paste(unique(gene_name), collapse = "; "),
    response_type = paste(unique(response_type), collapse = "; "),
    .groups = "drop"
  ) %>% arrange(response_type,gene_name)
select <- data.frame(select)
rownames(select) <- select$query

z_scores <- t(scale(t(assay(vsd)[select$query, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=FALSE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_row = (select %>% select(response_type)), annotation_colors = ann_colors,
         labels_row = select$gene_name, fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "HeatStress_all_SwissProt")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select$query,
  Heatmap_Label = select$gene_name,
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  annotation_row = (select %>% select(response_type)),
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  
  gaps_row = 4  # Add breaks only between clusters
)

# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/HeatStress_all_SwissProt")
``` 

### wnt and hox

```{r}
WNT <- Manual %>% filter(grepl("wnt", Heatmap_Label, ignore.case = TRUE)|grepl("frizzle", Heatmap_Label, ignore.case = TRUE)|grepl("homeobox", Heatmap_Label, ignore.case = TRUE)|grepl("hox", Heatmap_Label, ignore.case = TRUE))#|grepl("wnt", BiologicalProcess, ignore.case = TRUE))

select <- WNT$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "WNT_Hox_DE_SwissProt")

# Perform clustering
# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  gaps_row = 26  # Add breaks only between clusters
)


# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/WNT_Hox_DE_SwissProt")

```


#### wnt chromosome
```{r}
WNT_all <- Manual %>% filter(grepl("wnt", Heatmap_Label, ignore.case = TRUE)|grepl("frizzle", Heatmap_Label, ignore.case = TRUE)|grepl("homeobox", Heatmap_Label, ignore.case = TRUE)|grepl("hox", Heatmap_Label, ignore.case = TRUE))#|grepl("wnt", BiologicalProcess, ignore.case = TRUE))

WNT <- Manual %>% filter(grepl("wnt", Heatmap_Label, ignore.case = TRUE))
FRZ <- Manual %>% filter(grepl("frizzle", Heatmap_Label, ignore.case = TRUE))
HOX <- Manual %>% filter(grepl("hox", Heatmap_Label, ignore.case = TRUE)|grepl("homeobox", Heatmap_Label, ignore.case = TRUE))

select <- WNT$query

gene_layout <- gff_transcripts %>%
  filter(query %in% select) %>%
  arrange(chromosome, start) %>% left_join(WNT) %>% distinct() 

expr_df <- assay(vsd)[select, ] %>%
  as.data.frame() %>%
  rownames_to_column("query") %>%
  pivot_longer(-query, names_to = "Sample", values_to = "vsd_expr") %>%
  left_join(as.data.frame(colData(vsd)), by = "Sample") %>%
  group_by(query, Tissue) %>%
  summarise(mean_expr = mean(vsd_expr), .groups = "drop") %>%
  pivot_wider(names_from = Tissue, values_from = mean_expr)

plot_df <- gene_layout %>%
  left_join(expr_df, by = "query")

plot_df <- plot_df %>%
  mutate(highest_expr_tissue = case_when(
    OralEpi > Aboral ~ "OralEpi",
    Aboral > OralEpi ~ "Aboral",
    TRUE ~ "equal"
  ))

plot_df <- plot_df %>%
  mutate(chromosome_split = case_when(
    chromosome == "Pocillopora_acuta_HIv2___xfSc0000014" & start < 3.6e6 ~ "xfSc0000014_A",
    chromosome == "Pocillopora_acuta_HIv2___xfSc0000014" & start >= 3.65e6 ~ "xfSc0000014_B",
    TRUE ~ chromosome
  ))

plot_df$Heatmap_Label <- gsub("Protein ","",plot_df$Heatmap_Label)

heat_df <- plot_df %>%
  select(query, start, end, width,chromosome, OralEpi, Aboral) %>%
  pivot_longer(cols = c(OralEpi, Aboral), names_to = "tissue", values_to = "expression") %>%
  mutate(midpoint = (start + end) / 2)  # for positioning tile center

heat_df <- heat_df %>%
  left_join(plot_df %>% select(query, chromosome_split), by = "query") %>%
  mutate(chromosome = chromosome_split)

library(gggenes)
library(ggnewscale)

Wnt_chr <- ggplot(plot_df, aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_label(data = plot_df,
           aes(x = (start + end)/2, y = "_Genes_ ", label = Heatmap_Label),
           vjust = -1.75, size = 2, fill = "lightgrey", label.size = 0.1,fontface="bold")+
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome_split, scales = "free_x")+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")
Wnt_chr
save_ggplot(Wnt_chr, "Wnt_chromosome")
```


```{r}
Wnt_chr_heatmap <- ggplot() +
  # Heatmap tiles
  geom_tile(data = heat_df,
            aes(x = midpoint, y = tissue, fill = expression,width = width),colour="black",
            height = 0.5) +
  
  # Color for heatmap (expression level)
  scale_fill_gradient(low = "white", high = "firebrick", name = "Expression") +
  new_scale_fill() +
  
  # Gene arrows
  geom_gene_arrow(data = plot_df,
                  aes(xmin = start, xmax = end, y = "_Genes_",
                      forward = strand == "+", fill = highest_expr_tissue),
                  arrowhead_height = unit(3, "mm"),
                  arrowhead_width = unit(1.5, "mm")) +
    geom_label(data = plot_df,
           aes(x = (start + end)/2, y = "_Genes_", label = Heatmap_Label),
           vjust = 1.5, size = 4, fill = "lightgrey", label.size = 0.1,fontface="bold")+

  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey"),
                    name = "Higher expression in") +

  facet_wrap(~ chromosome_split, scales = "free_x") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_blank(),
    panel.spacing = unit(.1, "lines")
  ) +scale_y_discrete(expand = expansion(mult = c(1, 0)))+
  labs(x = "Genomic position")

save_ggplot(Wnt_chr_heatmap, "Wnt_chromosome_expression",width = 20, height = 5)
```

```{r}
# Shift gene coordinates within each chromosome
plot_df_shifted <- plot_df %>%
  group_by(chromosome) %>%
  mutate(
    chr_start = min(start),
    start_shifted = start - chr_start + 10,  # +10 if you want to avoid 0
    end_shifted   = end - chr_start + 10
  ) %>%
  ungroup()


heat_df_shifted <- plot_df_shifted %>%
  select(query, start_shifted, end_shifted, width,chromosome, OralEpi, Aboral) %>%
  pivot_longer(cols = c(OralEpi, Aboral), names_to = "tissue", values_to = "expression") %>%
  mutate(midpoint = (start_shifted + end_shifted) / 2)  # for positioning tile center

Wnt_chr_heatmap <- ggplot() +
  # Heatmap tiles
  geom_tile(data = heat_df_shifted,
            aes(x = midpoint, y = tissue, fill = expression,width = width),colour="black",
            height = 0.5) +
  
  # Color for heatmap (expression level)
  scale_fill_gradient(low = "white", high = "firebrick", name = "Expression") +
  new_scale_fill() +
  
  # Gene arrows
  geom_gene_arrow(data = plot_df_shifted,
                  aes(xmin = start_shifted, xmax = end_shifted, y = "_Genes_",
                      forward = strand == "+", fill = highest_expr_tissue),
                  arrowhead_height = unit(3, "mm"),
                  arrowhead_width = unit(1.5, "mm")) +
    geom_label(data = plot_df_shifted,
           aes(x = (start_shifted + end_shifted)/2, y = "_Genes_", label = Heatmap_Label),
           vjust = 1.5, size = 4, fill = "lightgrey", label.size = 0.1,fontface="bold")+

  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey"),
                    name = "Higher expression in") +

  facet_wrap(~ chromosome, scales = "fixed",ncol=1) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_blank(),
    panel.spacing = unit(.1, "lines")
  ) +scale_y_discrete(expand = expansion(mult = c(1, 0)))+
  labs(x = "Genomic position")

save_ggplot(Wnt_chr_heatmap, "Wnt_chromosome_expression_length_preserved",width = 30, height = 5)
```

#### Hox 

```{r}
HOX <- Manual %>% filter(grepl("homeobox", Heatmap_Label, ignore.case = TRUE)|grepl("hox", Heatmap_Label, ignore.case = TRUE))
select <- HOX$query

gene_layout <- gff_transcripts %>%
  filter(query %in% select) %>%
  arrange(chromosome, start) %>% left_join(HOX) %>% distinct() 

expr_df <- assay(vsd)[select, ] %>%
  as.data.frame() %>%
  rownames_to_column("query") %>%
  pivot_longer(-query, names_to = "Sample", values_to = "vsd_expr") %>%
  left_join(as.data.frame(colData(vsd)), by = "Sample") %>%
  group_by(query, Tissue) %>%
  summarise(mean_expr = mean(vsd_expr), .groups = "drop") %>%
  pivot_wider(names_from = Tissue, values_from = mean_expr)

plot_df <- gene_layout %>%
  left_join(expr_df, by = "query")

plot_df <- plot_df %>%
  mutate(highest_expr_tissue = case_when(
    OralEpi > Aboral ~ "OralEpi",
    Aboral > OralEpi ~ "Aboral",
    TRUE ~ "equal"
  ))

plot_df <- plot_df %>%
  mutate(chromosome_split = case_when(
    chromosome == "Pocillopora_acuta_HIv2___xfSc0000002" & start < 2.2e6 ~ "xfSc0000002_A",
    chromosome == "Pocillopora_acuta_HIv2___xfSc0000002" & start >= 6.4e6 ~ "xfSc0000002_B",
      chromosome == "Pocillopora_acuta_HIv2___Sc0000011" & start < 2e6 ~ "Sc0000011_A",
      chromosome == "Pocillopora_acuta_HIv2___Sc0000011" & start >= 2e6 ~ "Sc0000011_B",
    chromosome == "Pocillopora_acuta_HIv2___Sc0000024" & start < 420000 ~ "Sc0000024_A",
    chromosome == "Pocillopora_acuta_HIv2___Sc0000024" & start >= 2300000 ~ "Sc0000024_B",
          chromosome == "Pocillopora_acuta_HIv2___xfSc0000008" & start < 2e6 ~ "xfSc0000008_A",
      chromosome == "Pocillopora_acuta_HIv2___xfSc0000008" & start >= 2e6 ~ "xfSc0000008_B",
    TRUE ~ chromosome
  ))

heat_df <- plot_df %>%
  select(query, start, end, width,chromosome, OralEpi, Aboral) %>%
  pivot_longer(cols = c(OralEpi, Aboral), names_to = "tissue", values_to = "expression") %>%
  mutate(midpoint = (start + end) / 2)  # for positioning tile center

heat_df <- heat_df %>%
  left_join(plot_df %>% select(query, chromosome_split), by = "query") %>%
  mutate(chromosome = chromosome_split)

library(gggenes)
library(ggnewscale)

HOX_chr <- ggplot(plot_df[1:6,], aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_label(aes(x = (start + end)/2, y = "_Genes_ ", label = Heatmap_Label),
           vjust = -1.75, size = 2, fill = "lightgrey", label.size = 0.1,fontface="bold")+
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome_split, scales = "free_x",ncol=1)+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")
HOX_chr
 save_ggplot(HOX_chr, "Hox_chromosome")
 
 HOX_chr <- ggplot(plot_df[7:12,], aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_label(aes(x = (start + end)/2, y = "_Genes_ ", label = Heatmap_Label),
           vjust = -1.75, size = 2, fill = "lightgrey", label.size = 0.1,fontface="bold")+
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome_split, scales = "free_x",ncol=1)+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")
HOX_chr
 save_ggplot(HOX_chr, "Hox_chromosome2")
 
  HOX_chr <- ggplot(plot_df[13:19,], aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_label(aes(x = (start + end)/2, y = "_Genes_ ", label = Heatmap_Label),
           vjust = -1.75, size = 2, fill = "lightgrey", label.size = 0.1,fontface="bold")+
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome_split, scales = "free_x",ncol=1)+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")
HOX_chr
 save_ggplot(HOX_chr, "Hox_chromosome3")
 
   HOX_chr <- ggplot(plot_df[20:24,], aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_label(aes(x = (start + end)/2, y = "_Genes_ ", label = Heatmap_Label),
           vjust = -1.75, size = 2, fill = "lightgrey", label.size = 0.1,fontface="bold")+
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome_split, scales = "free_x",ncol=1)+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(2, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")
HOX_chr
 save_ggplot(HOX_chr, "Hox_chromosome4")
```

```{r}
HOX_chr_heatmap <- ggplot() +
  # Heatmap tiles
  geom_tile(data = heat_df,
            aes(x = midpoint, y = tissue, fill = expression,width = width),colour="black",
            height = 0.5) +
  
  # Color for heatmap (expression level)
  scale_fill_gradient(low = "white", high = "firebrick", name = "Expression") +
  new_scale_fill() +
  
  # Gene arrows
  geom_gene_arrow(data = plot_df,
                  aes(xmin = start, xmax = end, y = "_Genes_",
                      forward = strand == "+", fill = highest_expr_tissue),
                  arrowhead_height = unit(3, "mm"),
                  arrowhead_width = unit(1.5, "mm")) +
    geom_label(data = plot_df,
           aes(x = (start + end)/2, y = "_Genes_", label = Heatmap_Label),
           vjust = 1.5, size = 2 , fill = "lightgrey", label.size = 0.1,fontface="bold")+

  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey"),
                    name = "Higher expression in") +

  facet_wrap(~ chromosome_split, scales = "free_x") +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 5),
    axis.title.y = element_blank(),
    panel.spacing = unit(.1, "lines")
  ) +#scale_y_discrete(expand = expansion(mult = c(1, 0)))+
  labs(x = "Genomic position")

save_ggplot(HOX_chr_heatmap, "Hox_chromosome_expression",width = 20, height = 5)
```

```{r}
# Shift gene coordinates within each chromosome
plot_df_shifted <- plot_df %>%
  group_by(chromosome_split) %>%
  mutate(
    chr_start = min(start),
    start_shifted = start - chr_start + 10,  # +10 if you want to avoid 0
    end_shifted   = end - chr_start + 10
  ) %>%
  ungroup()


heat_df_shifted <- plot_df_shifted %>%
  select(query, start_shifted, end_shifted, width,chromosome_split, OralEpi, Aboral) %>%
  pivot_longer(cols = c(OralEpi, Aboral), names_to = "tissue", values_to = "expression") %>%
  mutate(midpoint = (start_shifted + end_shifted) / 2)  # for positioning tile center

HOX_chr_heatmap <- ggplot() +
  # Heatmap tiles
  geom_tile(data = heat_df_shifted,
            aes(x = midpoint, y = tissue, fill = expression,width = width),colour="black",
            height = 1) +
  
  # Color for heatmap (expression level)
  scale_fill_gradient(low = "white", high = "firebrick", name = "Expression") +
  new_scale_fill() +
  
  # Gene arrows
  geom_gene_arrow(data = plot_df_shifted,
                  aes(xmin = start_shifted, xmax = end_shifted, y = "_Genes_",
                      forward = strand == "+", fill = highest_expr_tissue),
                  arrowhead_height = unit(3, "mm"),
                  arrowhead_width = unit(1.5, "mm")) +
    geom_label(data = plot_df_shifted,
           aes(x = (start_shifted + end_shifted)/2, y = "_Genes_", label = Heatmap_Label),
           vjust = 1.5, size = 4, fill = "lightgrey", label.size = 0.1,fontface="bold")+

  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey"),
                    name = "Higher expression in") +

  facet_wrap(~ chromosome_split, scales = "fixed",ncol=1) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_blank(),
    panel.spacing = unit(.1, "lines")
  ) +scale_y_discrete(expand = expansion(mult = c(1, 0)))+
  labs(x = "Genomic position")

save_ggplot_big <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300) {
  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,limitsize = FALSE)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,limitsize = FALSE)
}

save_ggplot_big(HOX_chr_heatmap, "Hox_chromosome_expression_length_preserved",width = 15, height = 25)
```

### He et al 2023 TFs/genes of interest Nematostella (NOTE! I confirmed the DE_05 ones are all represented above )

#### differentially expressed
```{r}
select <- unique(DE_05_He_etal$query)

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "He_etal_Nvec_DE_SwissProt")

# Perform clustering
# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  
  gaps_row = 4  # Add breaks only between clusters
)

# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/He_etal_Nvec_DE_SwissProt")
```
#### all
```{r}
labels <- DESeq_He_etal$def_short
select <- DESeq_He_etal$query

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = labels, fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "He_etal_Nvec_all_SwissProt")

# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = labels,
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  # Optional border around clusters
  
  gaps_row = 9  # Add breaks only between clusters
)


# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/He_etal_Nvec_all_SwissProt")
```


```{r}
gene_layout <- gff_transcripts %>%
  filter(query %in% select) %>%
  arrange(chromosome, start) %>% left_join(DESeq_He_etal) %>% distinct() 

expr_df <- assay(vsd)[select, ] %>%
  as.data.frame() %>%
  rownames_to_column("query") %>%
  pivot_longer(-query, names_to = "Sample", values_to = "vsd_expr") %>%
  left_join(as.data.frame(colData(vsd)), by = "Sample") %>%
  group_by(query, Tissue) %>%
  summarise(mean_expr = mean(vsd_expr), .groups = "drop") %>%
  pivot_wider(names_from = Tissue, values_from = mean_expr)

plot_df <- gene_layout %>%
  left_join(expr_df, by = "query")

plot_df <- plot_df %>%
  mutate(highest_expr_tissue = case_when(
    OralEpi > Aboral ~ "OralEpi",
    Aboral > OralEpi ~ "Aboral",
    TRUE ~ "equal"
  ))

heat_df <- plot_df %>%
  select(query, start, end, width,chromosome, OralEpi, Aboral) %>%
  pivot_longer(cols = c(OralEpi, Aboral), names_to = "tissue", values_to = "expression") %>%
  mutate(midpoint = (start + end) / 2)  # for positioning tile center


library(gggenes)
library(ggnewscale)

Hox_chr <- ggplot(plot_df, aes(xmin = start, xmax = end, y = "Hox genes", forward = strand == "+",
                    fill = highest_expr_tissue)) +
  geom_gene_arrow(arrowhead_height = unit(3, "mm"), arrowhead_width = unit(1.5, "mm")) +
  geom_gene_label(aes(label = Gene_Name), size = 50) +
  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey")) +
  theme_minimal() +
  facet_wrap(~ chromosome, scales = "free_x", ncol = 1)+
  theme(axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        panel.spacing = unit(.1, "lines")) +
  labs(x = "Genomic position", fill = "Higher expression in")


Hox_chr_heatmap <- ggplot() +
  # Heatmap tiles
  geom_tile(data = heat_df,
            aes(x = midpoint, y = tissue, fill = expression),colour="black",
            height = 0.4) +
  
  # Color for heatmap (expression level)
  scale_fill_gradient(low = "white", high = "firebrick", name = "Expression") +
  new_scale_fill() +
  
  # Gene arrows
  geom_gene_arrow(data = plot_df,
                  aes(xmin = start, xmax = end, y = "_Genes_",
                      forward = strand == "+", fill = highest_expr_tissue),
                  arrowhead_height = unit(3, "mm"),
                  arrowhead_width = unit(1.5, "mm")) +

  geom_gene_label(data = plot_df,
                  aes(xmin = start, xmax = end, y = "_Genes_", label = Gene_Name),
                  size = 50) +

  scale_fill_manual(values = c("OralEpi" = "palegreen3", "Aboral" = "mediumpurple1", "equal" = "grey"),
                    name = "Higher expression in") +
  facet_wrap(~ chromosome, scales = "free_x", ncol = 1) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 10),
    axis.title.y = element_blank(),
    panel.spacing = unit(.1, "lines")
  ) +
  labs(x = "Genomic position")

save_ggplot_big(Hox_chr, "He_etal_Nvec_all_SwissProt_chromosome", width = 20, height = 40)
save_ggplot_big(Hox_chr_heatmap, "He_etal_Nvec_all_SwissProt_chromosome_expression", width = 20, height = 40)
```

### DuBuc et al 2018 Wnt/Aboral-Oral Patterning Genes, interest Nematostella

#### differentially expressed
```{r}
select <- unique(DE_05_DuBuc_etal$query)

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "DuBuc_etal_Nvec_DE_SwissProt")

# Perform clustering
# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white"
)

# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/DuBuc_etal_Nvec_DE_SwissProt")
```


### combine hox and WNT lists

```{r}
DESeq_He_etal$source <- "He_etal_2023"
DESeq_DuBuc_etal$source <- "DuBuc_etal_2018"

Hox_Literature <- rbind(DESeq_He_etal,DESeq_DuBuc_etal)

DESeq_Hox_Literature_collapsed <- Hox_Literature %>%
  group_by(query, baseMean, log2FoldChange, lfcSE, pvalue, padj) %>%
  summarize(
    Gene_Name = paste(unique(Gene_Name), collapse = "; "),
    Description = paste(unique(Description), collapse = "; "),
    def_short = paste(unique(def_short), collapse = "; "),
    source = paste(unique(source), collapse = "; "),
    .groups = "drop"
  )

DE_05_Hox_Literature_collapsed <- DESeq_Hox_Literature_collapsed %>% filter(padj<0.05) %>% filter(abs(log2FoldChange) >1)

WNT_HOX_MAN <- Manual %>% select(-X) %>% filter(grepl("wnt", Heatmap_Label, ignore.case = TRUE)|grepl("frizzle", Heatmap_Label, ignore.case = TRUE)|grepl("homeobox", Heatmap_Label, ignore.case = TRUE)|grepl("hox", Heatmap_Label, ignore.case = TRUE)) %>% select(-c(blast_hit,evalue,ProteinNames,BiologicalProcess,GeneOntologyIDs))# %>% rename(Description = Heatmap_Label)

COMBINED_HOX_WNT_all <-  full_join(DESeq_Hox_Literature_collapsed,WNT_HOX_MAN,by=c("query")) %>%
  mutate(
    baseMean = coalesce(baseMean.x, baseMean.y),
    log2FoldChange = coalesce(log2FoldChange.x, log2FoldChange.y),
    lfcSE = coalesce(lfcSE.x, lfcSE.y),
    pvalue = coalesce(pvalue.x, pvalue.y),
    padj = coalesce(padj.x, padj.y)
  ) %>%
  select(-ends_with(".x"), -ends_with(".y"))

COMBINED_HOX_WNT_DE05 <-  full_join(DE_05_Hox_Literature_collapsed,WNT_HOX_MAN,by=c("query")) %>%
  mutate(
    baseMean = coalesce(baseMean.x, baseMean.y),
    log2FoldChange = coalesce(log2FoldChange.x, log2FoldChange.y),
    lfcSE = coalesce(lfcSE.x, lfcSE.y),
    pvalue = coalesce(pvalue.x, pvalue.y),
    padj = coalesce(padj.x, padj.y)
  ) %>%
  select(-ends_with(".x"), -ends_with(".y"))
```


```{r}
select <- unique(COMBINED_HOX_WNT_DE05$query)

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = gene_labels[match(select,(gene_labels$query)),2], fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "Hox_All_DE")

# Perform clustering
# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = gene_labels[match(select, gene_labels$query), 2],
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  
  gaps_row = c(2,5,12,14,19,25) 
)

# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/Hox_All_DE")
```

```{r}
COMBINED_HOX_WNT_all <- COMBINED_HOX_WNT_all %>%  mutate(Heatmap_Label = coalesce(Heatmap_Label, def_short))
COMBINED_HOX_WNT_all$Heatmap_Label <- gsub("Protein Wnt", "Wnt", COMBINED_HOX_WNT_all$Heatmap_Label, ignore.case = TRUE)

select <- unique(COMBINED_HOX_WNT_all$query)

z_scores <- t(scale(t(assay(vsd)[select, ]), center = TRUE, scale = TRUE))
top50_DE <- pheatmap(z_scores, color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200), cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=TRUE, cutree_cols = 2,annotation_col=(heatmap_metadata%>% select(Tissue)), annotation_colors = ann_colors,
         labels_row = COMBINED_HOX_WNT_all$Heatmap_Label, fontsize_row = 8, cutree_rows = 2)
top50_DE
save_ggplot(top50_DE, "Hox_All")

# Perform clustering
# Perform clustering
row_clusters <- hclust(dist(z_scores))
cluster_assignments <- cutree(row_clusters, k = 2) # Adjust k for the number of clusters

# Create a dataframe for clustering and label management
clustered_data <- data.frame(
  query = select,
  Heatmap_Label = COMBINED_HOX_WNT_all$Heatmap_Label,
  cluster = cluster_assignments
)

# Reorder rows within each cluster alphabetically by their labels
clustered_data <- clustered_data %>%
  arrange(cluster, Heatmap_Label)

# Reorder the z_scores matrix and labels based on the new order
z_scores <- z_scores[match(clustered_data$query, rownames(z_scores)), ]
ordered_labels <- clustered_data$Heatmap_Label

# Ensure there is only 1 gap between clusters
#row_breaks <- which(cluster_assignments != cluster_assignments[1])  # Only where cluster switches

# Generate the heatmap with reordered rows and labels
top50_DE <- pheatmap(
  z_scores,
  color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
  cluster_rows = FALSE,  # Disable clustering since rows are pre-ordered
  cluster_cols = TRUE,
  cutree_cols = 2,
  annotation_col = (heatmap_metadata %>% select(Tissue)),
  annotation_colors = ann_colors,
  labels_row = ordered_labels,
  fontsize_row = 8,
  borders_color = "white",  
  gaps_row = c(2,5,16,18,19,24,33) 
)

# Save the heatmap
save_ggplot(top50_DE, "clusters_clean/Hox_All")
```

### random
```{r}
# Function to generate short names for proteins
generate_short_name <- function(data) {
  data %>%
    mutate(short_name = ifelse(nchar(ProteinNames) > 60, 
                               paste0(substr(ProteinNames, 1, 57), "..."), 
                               ProteinNames))
}

# Function to create gene labels
create_gene_labels <- function(data) {
  data %>%
    select(query, short_name) %>%
    mutate_all(~ ifelse(is.na(.), "", .)) # Replace NAs with ""
}

# Function to generate z-scores for selected genes
calculate_z_scores <- function(data, selection, vsd_matrix) {
  selected_genes <- match(selection, rownames(vsd_matrix))
  selected_genes <- selected_genes[!is.na(selected_genes)] # Remove NAs
  t(scale(t(assay(vsd_matrix)[selected_genes, ]), center = TRUE, scale = TRUE))
}

# Function to create and save heatmap
create_heatmap <- function(z_scores, labels_row, annotation_col, annotation_colors, filename) {
  heatmap <- pheatmap(z_scores, 
                      color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(200),
                      cluster_rows = TRUE, 
                      show_rownames = TRUE,
                      cluster_cols = TRUE, 
                      cutree_cols = 2,
                      annotation_col = annotation_col,
                      annotation_colors = annotation_colors,
                      labels_row = labels_row, 
                      fontsize_row = 6)
  save_ggplot(heatmap, filename)
}

# Process DE_05_SwissProt
DE_05_SwissProt <- generate_short_name(DE_05_SwissProt)
gene_labels <- create_gene_labels(DE_05_SwissProt)
GFP <- DE_05_SwissProt %>% filter(grepl("biolum", BiologicalProcess, ignore.case = TRUE))
select <- GFP$query
z_scores <- calculate_z_scores(DE_05_SwissProt, select, vsd)
create_heatmap(z_scores, 
               labels_row = gene_labels[match(select, gene_labels$query), 2], 
               annotation_col = heatmap_metadata %>% select(Tissue), 
               annotation_colors = ann_colors, 
               filename = "GFP_DE_SwissProt")

# Process annot_tab
annot_tab <- generate_short_name(annot_tab)
gene_labels <- create_gene_labels(annot_tab)
GFP <- annot_tab %>% filter(grepl("biolum", BiologicalProcess, ignore.case = TRUE))
select1 <- GFP$query
z_scores <- calculate_z_scores(annot_tab, select1, vsd)
create_heatmap(z_scores, 
               labels_row = gene_labels[match(select1, gene_labels$query), 2], 
               annotation_col = heatmap_metadata %>% select(Tissue), 
               annotation_colors = ann_colors, 
               filename = "GFP_SwissProt")
```


### Aboral Oral Larvae Paper - Pdam genome

#### Join our data to Pdam annotations

```{r load_annotation}
Annot_Pdam <- read.csv("../references/annotation/blastp_Pdam_out.tab", sep = '\t', header = FALSE) %>% select(c(1,2)) %>% dplyr::rename("protein_id" = "V2", "query" = "V1")
```

```{r}
Manual_Pdam <- left_join(Manual, Annot_Pdam) 
```

```{r}
library(readxl)
larval_aboral_enriched <- read_excel("../output_RNA/marker_genes/RamonMateu_Pdam_larval.xlsx", sheet = "pocillopora_aboral_enriched_05_")

Manual_Pdam_upaboral <- Manual_Pdam %>% filter(log2FoldChange > 0)
inner_join(larval_aboral_enriched, Manual_Pdam_upaboral, by=c("gene ID"="protein_id")) %>% dim()

#18 genes overlap from the 126 in the paper

larval_oral_enriched <- read_excel("../output_RNA/marker_genes/RamonMateu_Pdam_larval.xlsx", sheet = "pocillopora_oral_enriched_05_lf")

Manual_Pdam_uporal <- Manual_Pdam %>% filter(log2FoldChange < 0)
inner_join(larval_oral_enriched, Manual_Pdam_uporal, by=c("gene ID"="protein_id")) %>% dim()

inner_join(larval_oral_enriched, Manual_Pdam_uporal, by=c("gene ID"="protein_id")) %>% print()

#only 6/83 genes overlap


larval_aboral_clusters <- read_excel("../output_RNA/marker_genes/RamonMateu_Pdam_larval_scRNA.xlsx") %>% filter(!is.na(pocillopora))

larval_aboral_clusters_Pacuta <- inner_join(larval_aboral_clusters, Manual_Pdam, by=c("pocillopora"="protein_id"))

#my genes upregulated in oral tissue are high in mucous cells, cool
```

## Updating Renv environment:

After youve confirmed your code works as expected, use renv::snapshot() to record the packages and their sources in the lockfile.

```{r, eval=FALSE}
renv::snapshot()
```
