---
title: "DE"
author: "Zoe Dellaert"
date: "2024-10-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Renv

To run this code in my project using the renv environment, run the following lines of code

```{r}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Install packages

```{r}
renv::install("bioc::genefilter")
renv::install("DESeq2")
renv::install("tidyverse")

library(genefilter)
library(DESeq2)
library(tidyverse)
```

```{r}
sessionInfo() #provides list of loaded packages and version of R.
```

## Read and clean count matrix and metadata 

Read in raw count data
```{r}
counts_raw <- read.csv("../output_RNA/stringtie/LCM_RNA_gene_count_matrix.csv") #load in data
rownames(counts_raw) <- counts_raw[,1] #set first column that contains gene names as rownames
counts_raw <- counts_raw[,-1] #remove the column with gene names
```

gene_id,LCM_15,LCM_16,LCM_20,LCM_21,LCM_26,LCM_27,LCM_4,LCM_5,LCM_8,LCM_9

Read in metadata 
```{r}
meta <- read.csv("../data_RNA/LCM_RNA_metadata.csv")
meta <- dplyr::arrange(meta, Sample)

# Set variables as factors 
meta$Tissue <- factor(meta$Tissue, levels = c("OralEpi","Aboral")) #we want OralEpi to be the baseline

meta$Fragment <- factor(meta$Fragment)
meta$Section_Date <- factor(meta$Section_Date)
meta$LCM_Date <- factor(meta$LCM_Date)
```

Data sanity checks!
```{r}
all(meta$Sample %in% colnames(counts_raw)) #are all of the sample names in the metadata column names in the gene count matrix? Should be TRUE
all(meta$Sample == colnames(counts_raw)) #are they the same in the same order? Should be TRUE
```

## pOverA filtering to reduce dataset

```{r}
ffun<-filterfun(pOverA(0.25,5))  #set up filtering parameters
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter
sum(counts_filt_poa) #count number of genes left

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter
```

There are now 15530 genes in the filtered dataset.

Data sanity checks:
```{r}
all(meta$Sample %in% colnames(filtered_counts)) #are all of the sample names in the metadata column names in the gene count matrix?
all(meta$Sample == colnames(filtered_counts))  #are they the same in the same order? Should be TRUE
```

## Following [DESeq2 Vignette](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

Alternative filtering:

```{r}
smallestGroupSize <- 2
keep <- rowSums(counts_raw >= 10) >= smallestGroupSize
flitered_counts_2 <- counts_raw[keep,]
nrow(flitered_counts_2)
```

There are now 17546 genes in the filtered dataset.

Data sanity checks:
```{r}
all(meta$Sample %in% colnames(flitered_counts_2)) #are all of the sample names in the metadata column names in the gene count matrix?
all(meta$Sample == colnames(flitered_counts_2))  #are they the same in the same order? Should be TRUE
```

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ Fragment + Tissue)

dds <- DESeq(dds)
res <- results(dds)
res
dds <- DESeq(dds)
resultsNames(dds) # lists the coefficients
res <- results(dds, name="tissue_trt_vs_untrt")
# or to shrink log fold changes association with condition:
res <- lfcShrink(dds, coef="condition_trt_vs_untrt", type="apeglm")
```


## Updating Renv environment:

After you’ve confirmed your code works as expected, use renv::snapshot() to record the packages and their sources in the lockfile."
