---
title: "KEGG-Pathways"
author: "Zoe Dellaert"
date: "2024-12-09"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## KEGG Pathway analysis of LCM RNA Data

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r message=FALSE}
packages <- c("pathview", "KEGGREST", "rtracklayer" , "tidyverse")

lapply(packages, library, character.only = TRUE)
```

```{r}
sessionInfo() #provides list of loaded packages and version of R.
```

## Load in differential expression data

```{r}
DE_05 <- read.csv("../output_RNA/differential_expression/DE_05_SwissProt_annotation.csv", header = TRUE) %>% select(-X)
DESeq_allgenes <- read.csv("../output_RNA/differential_expression/DESeq_results.csv", header = TRUE) %>% dplyr::rename("query" ="X")

uniprot_ids <- DE_05$blast_hit[!is.na(DE_05$blast_hit)]
```

## BLAST Pacuta genes to Pdam genome

```{bash eval=FALSE}
cd ../scripts
nano Pdam_blast.sh
```

```{bash eval=FALSE}
#!/bin/bash
#SBATCH --job-name="Pdam_blast"
#SBATCH -t 240:00:00
#SBATCH --export=NONE
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails
#SBATCH --mail-user=zdellaert@uri.edu #your email to send notifications
#SBATCH --mem=100GB
#SBATCH --error=../scripts/outs_errs/"%x_error.%j" #write out slurm error reports
#SBATCH --output=../scripts/outs_errs/"%x_output.%j" #write out any program outpus
#SBATCH --account=putnamlab
#SBATCH --nodes=2 --ntasks-per-node=24

module load BLAST+/2.15.0-gompi-2023a

cd ../references

makeblastdb -in Pdam_RefSeq_protein.faa -out blast_dbs/Pdam_prot -dbtype prot

blastp -query Pocillopora_acuta_HIv2.genes.pep.faa -db blast_dbs/Pdam_prot -out annotation/blastp_Pdam_out.tab -outfmt 6 -evalue 1E-05 \
-num_threads 48 \
-max_target_seqs 1 \
-max_hsps 1

echo "Blast complete!" $(date)
```

```{bash, eval=FALSE}
sbatch Pdam_blast.sh
```

```{r}
Pdam_gtf <- import("../references/Pdam_RefSeq.gtf")
Pdam_gtf_proteins <- as.data.frame(Pdam_gtf) %>% filter(type == "CDS") %>% select(protein_id, gene_id) %>% distinct()

Annot_Pdam <- read.csv("../references/annotation/blastp_Pdam_out.tab", sep = '\t', header = FALSE) %>% select(c(1,2)) %>% dplyr::rename("protein_id" = "V2", "query" = "V1")

Annot_Pdam <- left_join(Annot_Pdam, Pdam_gtf_proteins)
```

Join to DE data:
```{r}
DE_05_Pdam <- left_join(DE_05, Annot_Pdam)

# make gene data matrix
gene_data <- DE_05_Pdam %>% select(gene_id,log2FoldChange) %>% drop_na() 
gene_data$gene_id <- gsub("LOC", "", gene_data$gene_id) %>% as.numeric()

gene.data <- setNames(gene_data$log2FoldChange, gene_data$gene_id)
```

What pathways are common in our DE genes?
```{r}
pdam_pathways <- keggLink("pathway", "pdam")
pdam_pathways <- data.frame(pdam_pathways, names(pdam_pathways)) 
pdam_pathways <- pdam_pathways %>% dplyr::rename("gene_id" = "names.pdam_pathways.")
pdam_pathways$gene_id <- gsub("pdam:", "", pdam_pathways$gene_id) %>% as.numeric()

DE_05_Pdam_pathways <- left_join(gene_data, pdam_pathways) %>%  drop_na() %>% count(pdam_pathways) %>% arrange(-n)

head(DE_05_Pdam_pathways)
```


Join to all genes data:
```{r}
allgenes_Pdam <- left_join(DESeq_allgenes, Annot_Pdam)

# make gene data matrix
gene_data_allgenes <- allgenes_Pdam %>% select(gene_id,log2FoldChange) %>% drop_na() 
gene_data_allgenes$gene_id <- gsub("LOC", "", gene_data_allgenes$gene_id) %>% as.numeric()

gene.data_allgenes <- setNames(gene_data_allgenes$log2FoldChange, gene_data_allgenes$gene_id)
```

## Wnt signaling pathway - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04310

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04310", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04310", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```

## mTOR signaling pathway - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04150

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04150", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04150", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```

## Neuroactive ligand-receptor interaction - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04080

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04080", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04080", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```
## Motor proteins - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04814

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04814", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04814", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```
## Lysosome - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04142

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04142", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04142", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```

## Cytoskeleton in muscle cells - Pocillopora damicornis (lace coral)

See pathway here: https://www.genome.jp/entry/pathway+pdam04820

```{r}
setwd("../output_RNA/differential_expression/KEGG")

pathview(gene.data = gene.data, pathway.id = "pdam04820", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
         out.suffix = "DE_05", 
         low = "mediumpurple1",
         high = "green")

pathview(gene.data = gene.data_allgenes, pathway.id = "pdam04820", species = "pdam", gene.idtype="KEGG", kegg.native = TRUE,
          out.suffix = "allgenes", 
          low = "mediumpurple1",
          high = "green")
```

## Updating Renv environment:

After you’ve confirmed your code works as expected, use renv::snapshot() to record the packages and their sources in the lockfile."
