---
title: "Enrichment"
author: "Zoe Dellaert"
date: "2024-12-09"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
always_allow_html: true
---

## Gene Ontology Analysis analysis of LCM RNA Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("ape")
require("scales")
require("tidyplots")
require("tidyverse")

sessionInfo() #provides list of loaded packages and version of R.
```

## Description of pipeline

I am going to perform functional enrichment of GO terms using [GO_MWU](https://github.com/z0on/GO_MWU).

As described in the [README](https://github.com/z0on/GO_MWU) of the package, the steps to running this analysis are as follows:

1. Make a directory for GO_MWU files. I am creating one in this code/ directory: LaserCoral/code/go_mwu
2. Download the following scripts into that directory: GO_MWU.R, gomwu_a.pl, gomwu_b.pl, gomwu.functions.R (downloaded from the github, linked above)
3. Download the GO hierarchy file, go.obo, from this website: http://www.geneontology.org/GO.downloads.ontology.shtml
4. table of GO annotations for your sequences:
  - "two-column (gene id - GO terms), tab-delimited, one line per gene, multiple GO terms separated by semicolon. If you have multiple lines per gene, use nrify_GOtable.pl to merge them. Do NOT include genes without GO annotations."
5. table of measure of interest for your sequences:
  - "two columns of comma-separated values: gene id, continuous measure of change such as log(fold-change). To perform standard GO enrichment analysis based on Fisher's exact test, use binary measure (1 or 0, i.e., either significant or not)."

## Load in reference files and differential expression data

In the next chunk I am loadinf in my DESeq data. These results are ordered by adjusted p-value. As a reminder, negative LFC = higher in Aboral tissue, and positive LFC = higher in Oral tissue.

```{r load_data}
#load in DESeq results
DESeq <- read.csv("../output_RNA/differential_expression/DESeq_results.csv", header = TRUE) %>% dplyr::rename("query" ="X")

#make dataframes of just differentially expressed genes for each LFC direction
DE_05_Aboral <- DESeq %>% filter(padj < 0.05 & log2FoldChange > 0)
DE_05_OralEpi <- DESeq %>% filter(padj < 0.05& log2FoldChange < 0)

#load in annotation data 
annot_tab <- read.delim("../references/annotation/protein-GO.tsv") %>% rename(GOs = GeneOntologyIDs)

#filter annotation data for just expressed genes 
annot_tab <- annot_tab %>% filter(query %in% DESeq$query) 

annot_tab$GOs <- gsub("; ", ";", annot_tab$GOs)
annot_tab$GOs[annot_tab$GOs==""] <- NA

nrow(annot_tab)
nrow(annot_tab)/nrow(DESeq)
```

10927/14464 genes in our dataset have annotation information in this file. That is 76%.

```{r}
sum(annot_tab$query %in% DE_05_Aboral$query)
sum(annot_tab$query %in% DE_05_Aboral$query)/nrow(DE_05_Aboral)

sum(annot_tab$query %in% DE_05_OralEpi$query)
sum(annot_tab$query %in% DE_05_OralEpi$query)/nrow(DE_05_OralEpi)
```

549/804 genes that are significantly upregulated in the Aboral tissue have annotation information. That is 68% of the genes.

2033/2802 genes that are significantly upregulated in the Oral Epidermis tissue have annotation information. That is 73% of the genes.

## Create gene_to_go.tab file for running GO_MWU

```{r}
GO.terms <- annot_tab %>% select(query,GOs) %>% dplyr::rename("GO.terms" = GOs)

GO.terms_only <- GO.terms %>% na.omit()

nrow(GO.terms_only)
nrow(GO.terms_only)/nrow(DESeq)

write.table(GO.terms_only, "go_mwu/gene_to_go.tab",row.names = FALSE, sep = "\t", quote = FALSE)
```

10638/14464 genes in our dataset have GO term information. That is 74%.

```{r}
sum(GO.terms_only$query %in% DE_05_Aboral$query)
sum(GO.terms_only$query %in% DE_05_Aboral$query)/nrow(DE_05_Aboral)

sum(GO.terms_only$query %in% DE_05_OralEpi$query)
sum(GO.terms_only$query %in% DE_05_OralEpi$query)/nrow(DE_05_OralEpi)
```

545/804 genes that are significantly upregulated in the Aboral tissue have GO term information. That is 68% of the genes.

1979/2802 genes that are significantly upregulated in the Oral Epidermis tissue have GO term information. That is 71% of the genes.

## Create the GO_MWU input csv files for the binary analysis

```{r}
### Generate vector with 1 for the 214 significant DEGs that have positive LFC and GO annotation, and 0 for the other genes that have GO annotation

UpAboral <- GO.terms_only %>% inner_join(DESeq) %>%
    mutate(DE = ifelse(padj < 0.05 & log2FoldChange > 0, 1, 0)) %>%
    dplyr::select(query, DE) 
  
print(sum(UpAboral$DE)) #only 545 of these have GO annotations
  
write.csv(UpAboral, "go_mwu/UpAboral.csv", row.names = FALSE, quote = FALSE)

### Generate vector with 1 for the 1045 significant DEGs that have positive LFC and GO annotation, and 0 for the other genes that have GO annotation

UpOralEpi <- GO.terms_only %>% inner_join(DESeq) %>%
    mutate(DE = ifelse(padj < 0.05 & log2FoldChange < 0, 1, 0)) %>%
    dplyr::select(query, DE) 
  
print(sum(UpOralEpi$DE)) #only 1979 of these have GO annotations
  
write.csv(UpOralEpi, "go_mwu/UpOralEpi.csv", row.names = FALSE, quote = FALSE)
```

## Create the GO_MWU input csv files for the continuous LFC analysis

```{r}
### Generate vector with all of the genes that are expressed and have GO annotation, with their Log2FoldChanges
DE_LFC <- GO.terms_only %>% inner_join(DESeq) %>%
    dplyr::select(query, log2FoldChange) 
  
write.csv(DE_LFC, "go_mwu/DE_LFC.csv", row.names = FALSE, quote = FALSE)
```

## Run GO_MWU

```{r}
setwd("go_mwu")

goAnnotations="gene_to_go.tab" # two-column, tab-delimited, one line per gene, multiple GO terms separated by semicolon.
goDatabase="go.obo" # download from http://www.geneontology.org/GO.downloads.ontology.shtml
goDivision="BP" # either MF, or BP, or CC
source("gomwu.functions.R")
```


### Up Aboral Fisher Exact Test

```{r, eval=FALSE}
setwd("go_mwu")
input="UpAboral.csv"

gomwuStats(input, goDatabase, goAnnotations, goDivision,
	perlPath="perl", # replace with full path to perl executable if it is not in your system's PATH already
	largest=0.1,  # a GO category will not be considered if it contains more than this fraction of the total number of genes
	smallest=5,   # a GO category should contain at least this many genes to be considered
	clusterCutHeight=0.25 # threshold for merging similar (gene-sharing) terms. See README for details.
#	,Alternative="g" # by default the MWU test is two-tailed; specify "g" or "l" of you want to test for "greater" or "less" instead.
)
```

```
go.obo gene_to_go.tab UpAboral.csv BP largest=0.1 smallest=5 cutHeight=0.25

Run parameters:

largest GO category as fraction of all genes (largest)  : 0.1
         smallest GO category as # of genes (smallest)  : 5
                clustering threshold (clusterCutHeight) : 0.25

-----------------
retrieving GO hierarchy, reformatting data...

-------------
go_reformat:
Genes with GO annotations, but not listed in measure table: 1

Terms without defined level (old ontology?..): 58
-------------
-------------
go_nrify:
12179 categories, 9535 genes; size range 5-953.5
	68 too broad
	6074 too small
	6037 remaining

removing redundancy:

calculating GO term similarities based on shared genes...
4441 non-redundant GO categories of good size
-------------

Secondary clustering:
calculating similarities....
Binary classification detected; will perform Fisher's test
185 GO terms at 10% FDR
```

#### Plotting

```{r}
setwd("go_mwu")
input="UpAboral.csv"

png(filename = paste0(input, "_plot.png"), width = 1200, height = 2400)
results=gomwuPlot(input,goAnnotations,goDivision,
 	absValue=0.001,  #Specify absValue=0.001 if you are doing Fisher's exact test for standard GO enrichment
 	level1=0.05, # FDR threshold for plotting. Specify level1=1 to plot all GO categories containing genes exceeding the absValue.
 	level2=0.01, # FDR cutoff to print in regular (not italic) font.
 	level3=0.001, # FDR cutoff to print in large bold font.
 	txtsize=2,    # decrease to fit more on one page, or increase (after rescaling the plot so the tree fits the text) for better "word cloud" effect
 	treeHeight=0.5, # height of the hierarchical clustering tree
 )
dev.off()
 
# text representation of results, with actual adjusted p-values
head(results[[1]])
```

#### extracting representative GOs

```{r}
# this module chooses GO terms that best represent *independent* groups of significant GO terms

pcut=1e-2 # adjusted pvalue cutoff for representative GO
hcut=0.9 # height at which cut the GO terms tree to get "independent groups". 

# plotting the GO tree with the cut level (un-remark the next two lines to plot)
 plot(results[[2]],cex=0.3)
 abline(h=hcut,col="red")

# cutting
ct=cutree(results[[2]],h=hcut)
annots=c();ci=1
for (ci in unique(ct)) {
  message(ci)
	rn=names(ct)[ct==ci]
	obs=grep("obsolete",rn)
	if(length(obs)>0) { rn=rn[-obs] }
	if (length(rn)==0) {next}
	rr=results[[1]][rn,]
	bestrr=rr[which(rr$pval==min(rr$pval)),]
	best=1
	if(nrow(bestrr)>1) {
		nns=sub(" .+","",row.names(bestrr))
		fr=c()
		for (i in 1:length(nns)) { fr=c(fr,eval(parse(text=nns[i]))) }
		best=which(fr==max(fr))
	}
	if (bestrr$pval[best]<=pcut) { annots=c(annots,sub("\\d+\\/\\d+ ","",row.names(bestrr)[best]))}
}

setwd("go_mwu")
mwus=read.table(paste("MWU",goDivision,input,sep="_"),header=T)
bestGOs=mwus[mwus$name %in% annots,]
bestGOs
```

---

### Oral Epidermis tissue fisher exact test

```{r, eval=FALSE}
setwd("go_mwu")
input="UpOralEpi.csv"

gomwuStats(input, goDatabase, goAnnotations, goDivision,
	perlPath="perl", # replace with full path to perl executable if it is not in your system's PATH already
	largest=0.1,  # a GO category will not be considered if it contains more than this fraction of the total number of genes
	smallest=5,   # a GO category should contain at least this many genes to be considered
	clusterCutHeight=0.25 # threshold for merging similar (gene-sharing) terms. See README for details.
#	,Alternative="g" # by default the MWU test is two-tailed; specify "g" or "l" of you want to test for "greater" or "less" instead.
)
```

```
go.obo gene_to_go.tab UpOralEpi.csv BP largest=0.1 smallest=5 cutHeight=0.25

Run parameters:

largest GO category as fraction of all genes (largest)  : 0.1
         smallest GO category as # of genes (smallest)  : 5
                clustering threshold (clusterCutHeight) : 0.25

-----------------
retrieving GO hierarchy, reformatting data...

-------------
go_reformat:
Genes with GO annotations, but not listed in measure table: 1

Terms without defined level (old ontology?..): 58
-------------
-------------
go_nrify:
12179 categories, 9535 genes; size range 5-953.5
	68 too broad
	6074 too small
	6037 remaining

removing redundancy:

calculating GO term similarities based on shared genes...
4441 non-redundant GO categories of good size
-------------

Secondary clustering:
calculating similarities....
Binary classification detected; will perform Fisher's test
17 GO terms at 10% FDR
```

#### Plotting

```{r}
setwd("go_mwu")
input="UpOralEpi.csv"

png(filename = paste0(input, "_plot.png"), width = 1200, height = 600)
results=gomwuPlot(input,goAnnotations,goDivision,
 	absValue=0.001,  #Specify absValue=0.001 if you are doing Fisher's exact test for standard GO enrichment
 	level1=0.05, # FDR threshold for plotting. Specify level1=1 to plot all GO categories containing genes exceeding the absValue.
 	level2=0.01, # FDR cutoff to print in regular (not italic) font.
 	level3=0.001, # FDR cutoff to print in large bold font.
 	txtsize=3,    # decrease to fit more on one page, or increase (after rescaling the plot so the tree fits the text) for better "word cloud" effect
 	treeHeight=0.25, # height of the hierarchical clustering tree
 )
dev.off()
 
# text representation of results, with actual adjusted p-values
head(results[[1]])
```

#### extracting representative GOs

```{r}
# this module chooses GO terms that best represent *independent* groups of significant GO terms

pcut=1e-2 # adjusted pvalue cutoff for representative GO
hcut=0.9 # height at which cut the GO terms tree to get "independent groups". 

# plotting the GO tree with the cut level (un-remark the next two lines to plot)
 plot(results[[2]],cex=0.3)
 abline(h=hcut,col="red")

# cutting
ct=cutree(results[[2]],h=hcut)
annots=c();ci=1
for (ci in unique(ct)) {
  message(ci)
	rn=names(ct)[ct==ci]
	obs=grep("obsolete",rn)
	if(length(obs)>0) { rn=rn[-obs] }
	if (length(rn)==0) {next}
	rr=results[[1]][rn,]
	bestrr=rr[which(rr$pval==min(rr$pval)),]
	best=1
	if(nrow(bestrr)>1) {
		nns=sub(" .+","",row.names(bestrr))
		fr=c()
		for (i in 1:length(nns)) { fr=c(fr,eval(parse(text=nns[i]))) }
		best=which(fr==max(fr))
	}
	if (bestrr$pval[best]<=pcut) { annots=c(annots,sub("\\d+\\/\\d+ ","",row.names(bestrr)[best]))}
}

setwd("go_mwu")
mwus=read.table(paste("MWU",goDivision,input,sep="_"),header=T)
bestGOs=mwus[mwus$name %in% annots,]
bestGOs
```

### Differential Expression using LFC

```{r, eval=FALSE}
setwd("go_mwu")
input="DE_LFC.csv"

gomwuStats(input, goDatabase, goAnnotations, goDivision,
	perlPath="perl", # replace with full path to perl executable if it is not in your system's PATH already
	largest=0.1,  # a GO category will not be considered if it contains more than this fraction of the total number of genes
	smallest=5,   # a GO category should contain at least this many genes to be considered
	clusterCutHeight=0.25 # threshold for merging similar (gene-sharing) terms. See README for details.
#	,Alternative="g" # by default the MWU test is two-tailed; specify "g" or "l" of you want to test for "greater" or "less" instead.
)
```

```
go.obo gene_to_go.tab DE_LFC.csv BP largest=0.1 smallest=5 cutHeight=0.25

Run parameters:

largest GO category as fraction of all genes (largest)  : 0.1
         smallest GO category as # of genes (smallest)  : 5
                clustering threshold (clusterCutHeight) : 0.25

-----------------
retrieving GO hierarchy, reformatting data...

-------------
go_reformat:
Genes with GO annotations, but not listed in measure table: 1

Terms without defined level (old ontology?..): 58
-------------
-------------
go_nrify:
12179 categories, 9535 genes; size range 5-953.5
	68 too broad
	6074 too small
	6037 remaining

removing redundancy:

calculating GO term similarities based on shared genes...
4441 non-redundant GO categories of good size
-------------

Secondary clustering:
calculating similarities....
Continuous measure of interest: will perform MWU test
78 GO terms at 10% FDR
```

#### Plotting

```{r}
setwd("go_mwu")
input="DE_LFC.csv"

png(filename = paste0(input, "_plot.png"), width = 1800, height = 1600)
results=gomwuPlot(input,goAnnotations,goDivision,
 	absValue=1, # un-remark this if you are using log2-fold changes
 	level1=0.05, # FDR threshold for plotting. Specify level1=1 to plot all GO categories containing genes exceeding the absValue.
 	level2=0.01, # FDR cutoff to print in regular (not italic) font.
 	level3=0.001, # FDR cutoff to print in large bold font.
 	txtsize=4.5,    # decrease to fit more on one page
 	treeHeight=.5, # height of the hierarchical clustering tree
 	# colors=c("dodgerblue2","firebrick1","skyblue2","lightcoral") 
 	colors = c("forestgreen", "purple", "seagreen3", "orchid")
 )
dev.off()
 
# text representation of results, with actual adjusted p-values
head(results[[1]])
```

#### extracting representative GOs

```{r}
# this module chooses GO terms that best represent *independent* groups of significant GO terms

pcut=1e-2 # adjusted pvalue cutoff for representative GO
hcut=0.9 # height at which cut the GO terms tree to get "independent groups". 

# plotting the GO tree with the cut level (un-remark the next two lines to plot)
 plot(results[[2]],cex=0.1)
 abline(h=hcut,col="red")

# cutting
ct=cutree(results[[2]],h=hcut)
annots=c();ci=1
for (ci in unique(ct)) {
  message(ci)
	rn=names(ct)[ct==ci]
	obs=grep("obsolete",rn)
	if(length(obs)>0) { rn=rn[-obs] }
	if (length(rn)==0) {next}
	rr=results[[1]][rn,]
	bestrr=rr[which(rr$pval==min(rr$pval)),]
	best=1
	if(nrow(bestrr)>1) {
		nns=sub(" .+","",row.names(bestrr))
		fr=c()
		for (i in 1:length(nns)) { fr=c(fr,eval(parse(text=nns[i]))) }
		best=which(fr==max(fr))
	}
	if (bestrr$pval[best]<=pcut) { annots=c(annots,sub("\\d+\\/\\d+ ","",row.names(bestrr)[best]))}
}

setwd("go_mwu")
mwus=read.table(paste("MWU",goDivision,input,sep="_"),header=T)
bestGOs=mwus[mwus$name %in% annots,]
bestGOs
```




## Custom plots

### up aboral, fisher

```{r}
MWU_BP_UpAboral <- read.table("go_mwu/MWU_BP_UpAboral.csv", header = TRUE)
BP_UpAboral <- read.table("go_mwu/BP_UpAboral.csv", header = TRUE)
```

```{r}
MWU_BP_UpAboral %>% dplyr::filter(p.adj<0.05) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = nseqs, fill = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x() %>%
  add_sum_bar() %>% #adjust_colors(new_colors = colors_continuous_mako) %>%
  sort_y_axis_labels(nseqs) %>%
  adjust_padding(right = 0.1) %>%
  add_data_labels(nseqs, label_position = c("right")) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, Up in Aboral Tissue, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_UpAboral_Fisher_05.png",
            width = 8, height = 15, units="in",bg = "transparent") 
```
### up oral epi, fisher

```{r}
MWU_BP_UpOralEpi <- read.table("go_mwu/MWU_BP_UpOralEpi.csv", header = TRUE)
BP_UpOralEpi <- read.table("go_mwu/BP_UpOralEpi.csv", header = TRUE)
```

```{r}
MWU_BP_UpOralEpi %>% dplyr::filter(p.adj<0.05) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = nseqs, fill = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x() %>%
  add_sum_bar() %>% #adjust_colors(new_colors = colors_continuous_mako) %>%
  sort_y_axis_labels(nseqs) %>%
  adjust_padding(right = 0.1) %>%
  add_data_labels(nseqs, label_position = c("right")) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, Up in Aboral Tissue, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_UpOralEpi_Fisher_05.png",
            width = 8, height = 5, units="in",bg = "transparent") 
```

### continuous by LFC

```{r}
MWU_BP_DE_LFC <- read.table("go_mwu/MWU_BP_DE_LFC.csv", header = TRUE)
BP_DE_LFC <- read.table("go_mwu/BP_DE_LFC.csv", header = TRUE)
```

```{r}
MWU_BP_DE_LFC %>% dplyr::filter(p.adj<0.05 & delta.rank > 0) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = nseqs, fill = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x(fontsize=9) %>%
  add_sum_bar() %>% adjust_colors(new_colors = colors_continuous_turbo) %>%
  sort_y_axis_labels(delta.rank) %>%
  adjust_padding(right = 0.1) %>%
  add_data_labels(delta.rank, label_position = c("right")) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, Up in Aboral Tissue, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_UpAboral.png",
            width = 8, height = 6, units="in",bg = "transparent") 

MWU_BP_DE_LFC %>% dplyr::filter(p.adj<0.05 & delta.rank < 0) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = nseqs, fill = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x(fontsize=9) %>%
  add_sum_bar() %>% adjust_colors(new_colors = colors_continuous_turbo) %>%
  sort_y_axis_labels(-delta.rank) %>%
  adjust_padding(right = 0.1) %>%
  add_data_labels(delta.rank, label_position = c("right")) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, Up in Oral Epidermis Tissue, padj < 0.05")  %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_UpOralEpi.png",
            width = 8, height = 1.75, units="in",bg = "transparent")
```

```{r}
MWU_BP_DE_LFC %>% dplyr::filter(p.adj<0.05) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = delta.rank, color = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x(fontsize=8) %>%
  add_sum_bar() %>% #add_sum_dot() %>% 
  adjust_colors(new_colors = colors_continuous_turbo) %>%
  sort_y_axis_labels(delta.rank) %>%
  adjust_padding(right = 0.1, left = 0.1) %>%
  #add_data_labels(delta.rank, label_position = c("right")) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU.png",
            width = 8, height = 6, units="in",bg = "transparent")
            
            
MWU_BP_DE_LFC %>% dplyr::filter(p.adj<0.05) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = delta.rank) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x(fontsize=8) %>%
  add_sum_bar(width=0.2, data = filter_rows(delta.rank < 0), fill = "palegreen4") %>%
  add_sum_bar(width=0.2, data = filter_rows(delta.rank > 0), fill = "mediumpurple") %>%
  add(ggplot2::stat_summary(data = filter_rows(delta.rank < 0),fun = identity,geom = "point",
                            aes(size = log10padj), color="palegreen4", alpha=0.5)) %>%
  add(ggplot2::stat_summary(data = filter_rows(delta.rank > 0),fun = identity,geom = "point",
                            aes(size = log10padj), color="mediumpurple", alpha=0.5)) %>%
  sort_y_axis_labels(delta.rank) %>%
  adjust_padding(right = 0.1, left = 0.1) %>%
  #add(ggplot2::scale_size_continuous(name = "$-log10~(Adjusted~p-value)$", range=c(0,6))) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_dots.png",
            width = 8, height = 6, units="in",bg = "transparent")
```

```{r}
MWU_BP_DE_LFC %>% dplyr::filter(p.adj<0.05) %>% 
  mutate(log10padj = -log10(p.adj)) %>%
  tidyplot(y = name, x = delta.rank, color = log10padj) %>%
  adjust_size(height = NA, width=NA) %>%
  theme_minimal_x(fontsize=8) %>%
  add_sum_bar(width=0.2, fill="gray") %>% 
  add(ggplot2::stat_summary(fun = identity,geom = "point",
                            aes(size = nseqs), alpha=0.8)) %>%
  adjust_colors(new_colors = c("#0000FF", "#CA0088", "#FF0000")) %>%
  sort_y_axis_labels(delta.rank) %>%
  adjust_padding(right = 0.1, left = 0.1) %>%
  adjust_legend_title("$-log10~(Adjusted~p-value)$") %>%
  add_title("Enriched GO terms by GO_MWU, padj < 0.05") %>%
  save_plot("../output_RNA/differential_expression/enrichment/GO_MWU_bubble.png",
            width = 8, height = 6, units="in",bg = "transparent")
```

