---
title: "Semantic Enrichment"
author: "Zoe Dellaert"
date: "2025-01-11"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
always_allow_html: true
---

## Gene Ontology Analysis analysis of LCM RNA Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("ViSEAGO")
require("topGO")
require("tidyverse")

sessionInfo() #provides list of loaded packages and version of R.
```

## Description of pipeline

I am going to perform functional enrichment of GO terms using [ViSEAGO](https://biodatamining.biomedcentral.com/articles/10.1186/s13040-019-0204-1). 

I am following this vignette: [http://bioconductor.unipi.it/packages/devel/bioc/vignettes/ViSEAGO/inst/doc/ViSEAGO.html](http://bioconductor.unipi.it/packages/devel/bioc/vignettes/ViSEAGO/inst/doc/ViSEAGO.html).

## Load in reference files and differential expression data

In the next chunk I am loading in my DESeq data. These results are ordered by adjusted p-value. As a reminder, negative LFC = higher in Aboral tissue, and positive LFC = higher in Oral tissue.

```{r load_data}
#load in DESeq results
DESeq <- read.csv("../output_RNA/differential_expression/DESeq_results.csv", header = TRUE) %>% dplyr::rename("query" ="X")

#make dataframes of just differentially expressed genes for each LFC direction
DE_05_Aboral <- DESeq %>% filter(padj < 0.05 & log2FoldChange > 0)
DE_05_OralEpi <- DESeq %>% filter(padj < 0.05& log2FoldChange < 0)

#load in annotation data 
annot_tab <- read.delim("../references/annotation/protein-GO.tsv") %>% dplyr::rename(GOs = GeneOntologyIDs)

#filter annotation data for just expressed genes with GO annotations
annot_tab <- annot_tab %>% filter(query %in% DESeq$query) 

annot_tab$GOs <- gsub("; ", ";", annot_tab$GOs)
annot_tab$GOs[annot_tab$GOs==""] <- NA
annot_tab <- annot_tab %>% filter(!is.na(GOs))

nrow(annot_tab)
nrow(annot_tab)/nrow(DESeq)
```

10638/14464 genes in our dataset have GO information in this file. That is 74%.

```{r}
sum(annot_tab$query %in% DE_05_Aboral$query)
sum(annot_tab$query %in% DE_05_Aboral$query)/nrow(DE_05_Aboral)

sum(annot_tab$query %in% DE_05_OralEpi$query)
sum(annot_tab$query %in% DE_05_OralEpi$query)/nrow(DE_05_OralEpi)
```

545/804 genes that are significantly upregulated in the Aboral tissue have annotation information. That is 68% of the genes.

1979/2802 genes that are significantly upregulated in the Oral Epidermis tissue have annotation information. That is 71% of the genes.

## Create custom GO annotation file for ViSEAGO

```{r}
##Get a list of GO Terms for all genes
annots <- annot_tab %>% dplyr::select(query,GOs) %>% dplyr::rename("GO.terms" = GOs)

# format into the format required by ViSEAGO for custom mappings
Custom_GOs <- annots %>%
  # Separate GO terms into individual rows
  separate_rows(GO.terms, sep = ";") %>%
  # Add necessary columns
  mutate(
    taxid = "pacuta",
    gene_symbol = query,
    evidence = "SwissProt"
  ) %>%
  # Rename columns
  dplyr::rename(
    gene_id = query,
    GOID = GO.terms
  ) %>%
  dplyr::select(taxid, gene_id, gene_symbol, GOID, evidence)

Custom_GOs_valid <- Custom_GOs %>% filter(GOID %in% keys(GO.db))

write.table(Custom_GOs_valid, "../output_RNA/differential_expression/semantic-enrichment/custom_GOs.txt",row.names = FALSE, sep = "\t", quote = FALSE,col.names=TRUE)

length(unique(Custom_GOs$gene_id))
length(unique(Custom_GOs_valid$gene_id))
```

We seem to have lost one gene when filtering for valid GO terms, so I need to account for that below.

### load the file into ViSEAGO

```{r}
Custom_Pacuta <- ViSEAGO::Custom2GO("../output_RNA/differential_expression/semantic-enrichment/custom_GOs.txt")

myGENE2GO_Pacuta <- ViSEAGO::annotate(
    id="pacuta",
    Custom_Pacuta
)
```

## Create gene lists for enrichment

```{r}
selection <- DESeq %>% filter(query %in% Custom_GOs_valid$gene_id) %>% 
                       mutate(DE_05_Aboral = ifelse(query %in% DE_05_Aboral$query, 1,0)) %>%
                       mutate(DE_05_Oral = ifelse(query %in% DE_05_OralEpi$query, 1,0)) %>%
                       mutate(expressed = 1)

selection_Aboral <- selection %>% pull(DE_05_Aboral) %>% as.factor()
names(selection_Aboral) <- selection %>% pull(query)

selection_Oral <- selection %>% pull(DE_05_Oral) %>% as.factor()
names(selection_Oral) <- selection %>% pull(query)

expressed <- selection %>% pull(expressed) %>% as.factor()
names(expressed) <- selection %>% pull(query)
```

## create topGO objects and perform enrichment using topGO wrapped by ViSEAGO

```{r}
#create topgo object using custom mappings
# BP_Aboral <- new("topGOdata",
#               ontology = "BP",
#               allGenes = selection_Aboral,
#               nodeSize = 5,
#               gene2GO = geneID2GO,
#               annot = annFUN.gene2GO) 

# create viseago object
selection <- names(selection_Aboral)[selection_Aboral==1]
background <- names(expressed)

BP_Aboral <- ViSEAGO::create_topGOdata(
    geneSel=selection,
    allGenes=background,
    gene2GO=myGENE2GO_Pacuta, 
    ont="BP",
    nodeSize=5
)

# perform TopGO test using classic algorithm
classic_Aboral <- topGO::runTest(
    BP_Aboral,
    algorithm ="classic",
    statistic = "fisher"
)

#create topgo object using custom mappings
# BP_Oral <- new("topGOdata",
#               ontology = "BP",
#               allGenes = selection_Oral,
#               nodeSize = 5,
#               gene2GO = geneID2GO,
#               annot = annFUN.gene2GO) 

# create viseago object
selection <- names(selection_Oral)[selection_Oral==1]
background <- names(expressed)

BP_Oral <- ViSEAGO::create_topGOdata(
    geneSel=selection,
    allGenes=background,
    gene2GO=myGENE2GO_Pacuta, 
    ont="BP",
    nodeSize=5
)

# perform TopGO test using classic algorithm
classic_Oral <- topGO::runTest(
    BP_Oral,
    algorithm ="classic",
    statistic = "fisher"
)
```

### Merge results for both conditions

```{r}
BP_sResults <- ViSEAGO::merge_enrich_terms(
    Input = list(
        Aboral = c("BP_Aboral", "classic_Aboral"),
        Oral = c("BP_Oral", "classic_Oral")
    )
)
```

### Visualize and save initial results

```{r}
# display the merged table
ViSEAGO::show_table(BP_sResults)

# print the merged table in a file
ViSEAGO::show_table(
    BP_sResults,
    "../output_RNA/differential_expression/semantic-enrichment/BP_sResults.csv"
)
```


```{r}
# count significant (or not) pvalues by condition
ViSEAGO::GOcount(BP_sResults)

# display interactions
ViSEAGO::Upset(
    BP_sResults,
    file="../output_RNA/differential_expression/semantic-enrichment/OLexport.csv"
)
```

## Semantic similarity

```{r}
# initialize 
myGOs<-ViSEAGO::build_GO_SS(
    gene2GO=myGENE2GO_Pacuta,
    enrich_GO_terms=BP_sResults
)

# compute all available Semantic Similarity (SS) measures
myGOs <- ViSEAGO::compute_SS_distances(
    myGOs,
    distance="Wang"
)
```

## Visualization

### Multi Dimensional Scaling

```{r}
# display MDSplot
ViSEAGO::MDSplot(myGOs)
```

### Heatmap

```{r}
# GOterms heatmap with the default parameters
Wang_clusters_wardD2<-ViSEAGO::GOterms_heatmap(
    myGOs,
    showIC=TRUE,
    showGOlabels=TRUE,
    GO.tree=list(
        tree=list(
            distance="Wang",
            aggreg.method="ward.D2"
        ),
        cut=list(
            dynamic=list(
                pamStage=TRUE,
                pamRespectsDendro=TRUE,
                deepSplit=2,
                minClusterSize =2
            )
        )
    ),
    samples.tree=NULL
)
```

```{r}
# Display the clusters-heatmap
ViSEAGO::show_heatmap(
    Wang_clusters_wardD2,
    "GOterms"
)
```


```{r}
# Display the clusters-heatmap table
ViSEAGO::show_table(Wang_clusters_wardD2)

# Print the clusters-heatmap table
ViSEAGO::show_table(
    Wang_clusters_wardD2,
    "../output_RNA/differential_expression/semantic-enrichment/cluster_heatmap_Wang_wardD2.csv"
)
```

### MDS

```{r}
# display colored MDSplot
ViSEAGO::MDSplot(
    Wang_clusters_wardD2,
    "GOterms"
)
```

## Visualization and interpretation of GO clusters

```{r}
# calculate semantic similarites between clusters of GO terms
Wang_clusters_wardD2<-ViSEAGO::compute_SS_distances(
    Wang_clusters_wardD2,
    distance=c("max", "avg","rcmax", "BMA")
)
```

```{r}
# build and highlight in an interactive MDSplot grouped clusters for one distance object
ViSEAGO::MDSplot(
    Wang_clusters_wardD2,
    "GOclusters"
)
```

```{r}
# GOclusters heatmap
Wang_clusters_wardD2<-ViSEAGO::GOclusters_heatmap(
    Wang_clusters_wardD2,
    tree=list(
        distance="BMA",
        aggreg.method="ward.D2"
    )
)
```


```{r}
# display the GOClusters heatmap
ViSEAGO::show_heatmap(
    Wang_clusters_wardD2,
    "GOclusters"
)
```


