---
title: "09-MethylKit, filtered for >90% conversion efficiency"
author: "Zoe Dellaert"
date: "2025-02-09"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
always_allow_html: true
---

## MethylKit - Reads filtered for >90% conversion efficiency

I am identifying differentially methylated loci using methylkit based on [Yaamini Venkataraman's code](https://osf.io/u46xj)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("methylKit")
require("parallel")
require("tidyverse")
require("vegan")
require("gplots")
require("ggplot2")
require("ggrepel")
require("ggpmisc")
require("dichromat")
require("readr")
require("genomationData")
require("genomation")

sessionInfo() #provides list of loaded packages and version of R.
```


```{r}
save_ggplot <- function(plot, filename, width = 7, height = 5, units = "in", dpi = 300) {
  # Display plot
  print(plot)
  
  # Save plot
  ggsave(filename = paste0(filename, ".png"), plot = plot, width = width, height = height, units = units, dpi = dpi)
}
```

```{r}
meta <- read.csv("../data_WGBS/LCM_WGBS_metadata.csv", sep = ",", header = TRUE) %>%
  mutate(Section_Date = as.character(Section_Date), LCM_Date = as.character(LCM_Date),DNA_Extraction_Date = as.character(DNA_Extraction_Date))

meta <- meta %>% arrange(Sample)
```

```{r}
file_list <- list.files("../output_WGBS/methylseq_V3_bwa_test/methyldackel/min_efficiency_test_new",pattern = "^min_90.*_CpG.methylKit$",  full.names = TRUE, include.dirs = FALSE)

sample <- gsub("_CpG.methylKit", "", basename(file_list) )
sample <- gsub("min_90_", "", sample)

sample == meta$Sample #the files and metadata are in the same order

tissue <- meta$Tissue
tissue_binary <- gsub("Aboral", "1", tissue)
tissue_binary <- gsub("OralEpi", "0", tissue_binary)
tissue_binary <- as.numeric(tissue_binary)
fragment <- meta$Fragment
```

```{r,eval=FALSE}
# methylObj=methRead(as.list(file_list),
#            sample.id = as.list(sample),
#            assembly = "Pacuta",
#            treatment = tissue_binary,
#            context = "CpG",
#            mincov = 5
#            )

# save(methylObj, file = "../output_WGBS/MethylKit.RData")

methyl_list <- mclapply(seq_along(file_list), function(i) {
  methRead(
    file_list[[i]],
    sample.id = sample[[i]],
    assembly = "Pacuta",
    treatment = tissue_binary[[i]],
    context = "CpG",
    mincov = 5
  )
}, mc.cores = 4) 

methylObj <- new("methylRawList", methyl_list, treatment = tissue_binary)

save(methylObj, file = "../output_WGBS/MethylKit_20250513.RData")
```


```{r}
load("../output_WGBS/MethylKit_20250513.RData")

getMethylationStats(methylObj[[2]],plot=FALSE,both.strands=FALSE)
getMethylationStats(methylObj[[2]],plot=TRUE,both.strands=FALSE)
getCoverageStats(methylObj[[2]],plot=TRUE,both.strands=FALSE)
getCoverageStats(methylObj[[2]],plot=TRUE,both.strands=TRUE)
```



### Note: I changed the code below so that a CpG does not have to have 5X coverage in all samples to be analyzed, and only needs 5X coverage in 2 samples per group to be retained. This way we don't remove CpGs that happen to have lower coverage in a few samples or one tissue.

```{r}
#filtered_methylObj=filterByCoverage(methylObj,lo.count=5,lo.perc=NULL,
#                                      hi.count=NULL,hi.perc=99.9)

#filtered_methylObj_norm <- filtered_methylObj %>% methylKit::normalizeCoverage(.)
methylObj_norm <- methylObj %>% methylKit::normalizeCoverage(.)
```

```{r}
#meth_filter=methylKit::unite(filtered_methylObj_norm)
meth_filter=methylKit::unite(methylObj_norm, min.per.group = c(3L,3L))

#meth_filter_destrand=methylKit::unite(filtered_methylObj_norm,destrand = TRUE)
meth_filter_destrand=methylKit::unite(methylObj_norm, min.per.group = c(3L,3L), destrand = TRUE)
 
clusterSamples(meth_filter, dist="correlation", method="ward", plot=TRUE)
clusterSamples(meth_filter_destrand, dist="correlation", method="ward", plot=TRUE)

PCASamples(meth_filter)
PCASamples(meth_filter_destrand)
```

```{r}
getCorrelation(meth_filter_destrand,plot=TRUE)
```

### batch effects 

```{r}
#as=assocComp(mBase=meth_filter_destrand,dplyr::select(meta,c("PCR_ReAmp_Cycles", "Fragment")))
#as
```


### other possible filtering

```{r}
# get percent methylation matrix
pm=percMethylation(meth_filter_destrand)

# calculate standard deviation of CpGs
sds=matrixStats::rowSds(pm)

# Visualize the distribution of the per-CpG standard deviation
# to determine a suitable cutoff
hist(sds, breaks = 100)

# keep only CpG with standard deviations larger than 2%
#meth <- meth_filter_destrand[sds > 2]

# This leaves us with this number of CpG sites
nrow(meth_filter_destrand)
#nrow(meth)
```


### Identify DML

```{r}
DMLStats_Tissue <- methylKit::calculateDiffMeth(meth_filter_destrand, overdispersion = "MN", test = "Chisq", mc.cores = 8) #Calculate differential methylation statistics and include covariate information.

head(DMLStats_Tissue) #Look at differential methylation output
```

```{r}
# Filter DMRs with q-value < 0.05
significant_dmg <- getData(DMLStats_Tissue[DMLStats_Tissue$qvalue < 0.05, ])

# Create a data frame for plotting
plot_data <- data.frame(
  chr = significant_dmg$chr,
  start = significant_dmg$start,
  meth.diff = significant_dmg$meth.diff
)

# Count the number of positive and negative methylation differences
positive_count <- sum(significant_dmg$meth.diff > 0)
negative_count <- sum(significant_dmg$meth.diff < 0)

# Plot with counts added to the quadrants
DML_direction <- ggplot(plot_data, aes(x = start, y = meth.diff)) +
  geom_point(alpha = 0.5) +  # Set alpha to reduce point transparency
  theme_minimal() +
  labs(title = "Significant Differentially Methylated Regions (q-value < 0.05)",
       x = "Genomic Position (start)",
       y = "Methylation Difference (%)") +
  theme(legend.position = "none") +  # Remove the legend
  # Add the count of positive and negative methylation differences as text annotations
  annotate("text", x = Inf, y = Inf, label = paste("Positive:", positive_count), 
           hjust = 1.1, vjust = 1.1, size = 4, color = "blue") +
  annotate("text", x = Inf, y = -Inf, label = paste("Negative:", negative_count), 
           hjust = 1.1, vjust = -0.1, size = 4, color = "red")

save_ggplot(DML_direction, "../output_WGBS/figures/6_DML_direction")


dml_df <- as.data.frame(DMLStats_Tissue)

# Volcano plot
DML_volcano <- ggplot(dml_df, aes(x = meth.diff, y = -log10(qvalue))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = c(-5, 5), linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(title = "Differentially Methylated Loci (DMLs)",
       x = "Methylation Difference (%)",
       y = "-log10(q-value)") +
  theme_minimal()

save_ggplot(DML_volcano, "../output_WGBS/figures/7_DML_volcano")

```


```{r}
DMLs <- methylKit::getMethylDiff(DMLStats_Tissue, difference = 2, qvalue = 0.05) #Identify DML based on difference threshold

length(DMLs$chr) #DML

head(DMLs)

DMLs_25 <- methylKit::getMethylDiff(DMLStats_Tissue, difference = 25, qvalue = 0.05) #Identify DML based on difference threshold

length(DMLs_25$chr) #DML

head(DMLs_25)
```

## Further look at genome wide methylation

```{r}
diffMethPerChr(DMLStats_Tissue, meth.cutoff = 2, qvalue.cutoff = 0.05,cex.names=.75)
```


### Annotation

```{r}
require("genomationData")
require("genomation")

gff.file = "../references/Pocillopora_acuta_HIv2.gtf"
gff = gffToGRanges(gff.file)
head(gff)
```


```{r}
exons = gffToGRanges(gff.file, filter = "exon")

transcripts = gffToGRanges(gff.file, filter = "transcript")

DML_all_grange = as(DMLStats_Tissue,"GRanges")
DML_grange = as(DMLs,"GRanges")
DML_grange
```


```{r}
transcripts = gffToGRanges(gff.file, filter = "transcript")

# Find overlaps between DMLs and transcripts
overlaps_transcripts <- findOverlaps(DML_all_grange, transcripts,ignore.strand = TRUE)

# Extract matching transcript information
DML_transcript_annot_allCpg <- data.frame(
  chr = seqnames(DML_all_grange)[queryHits(overlaps_transcripts)],
  start = start(DML_all_grange)[queryHits(overlaps_transcripts)],
  end = end(DML_all_grange)[queryHits(overlaps_transcripts)],
  DML_qvalue = (DML_all_grange$qvalue)[queryHits(overlaps_transcripts)],
  DML_methdiff = (DML_all_grange$meth.diff)[queryHits(overlaps_transcripts)],
  transcript_chr = seqnames(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_start = start(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_end = end(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_id = transcripts$transcript_id[subjectHits(overlaps_transcripts)],
  gene_id = transcripts$gene_id[subjectHits(overlaps_transcripts)]
)

DML_transcript_annot <- DML_transcript_annot_allCpg %>% filter(DML_qvalue < 0.05)
  
# View first few rows
head(DML_transcript_annot)

# how many DMLs are in gene bodies?
length(DML_transcript_annot$gene_id)

#how many genes do these consist of?
length(unique(DML_transcript_annot$gene_id))

DML_transcript_annot %>% group_by(transcript_id) %>% summarize(num_DMLs=n()) %>% summary()
```

## Are any DMGs DMLs?

```{r}
#load in DESeq results
DESeq <- read.csv("../output_RNA/differential_expression/DESeq_results.csv", header = TRUE) %>% dplyr::rename("query" ="X")

#make dataframes of just differentially expressed genes for each LFC direction
DE_05_Aboral <- DESeq %>% filter(padj < 0.05 & log2FoldChange > 0)
DE_05_OralEpi <- DESeq %>% filter(padj < 0.05& log2FoldChange < 0)
DE_05 <- DESeq %>% filter(padj < 0.05)

DML_transcript_annot[DML_transcript_annot$transcript_id %in% DE_05$query,]

DE_05[DE_05$query %in% DML_transcript_annot$transcript_id,]
```


```{r}
plot_data <- merge(DML_transcript_annot, DESeq, by.x = "transcript_id", by.y = "query")

ggplot(plot_data, aes(x = DML_methdiff, y = log2FoldChange)) +
  geom_point(aes(color = padj < 0.05), size = 3) +
  geom_text_repel(aes(label = ifelse(padj < 0.05, transcript_id, "")), max.overlaps = 20) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("grey", "red")) +
  labs(x = "Methylation Difference (%)", 
       y = "RNA-seq: Log2FoldChange",
       title = "Differentially methylated loci (DMLs) vs. Gene Expression",
       color = "Significant DEGs") +
  theme_minimal()
```

```{r}
# how many CpG sites are there after filtering?
dim(meth_filter_destrand)

CpG_grange = as(meth_filter_destrand,"GRanges")

# identify all the CpGs that are in transcripts
genes_with_CpGs <- findOverlaps(CpG_grange, transcripts,ignore.strand = TRUE)

# Extract matching transcript information
CpG_transcript_annot <- data.frame(
  CpG_chr = seqnames(CpG_grange)[queryHits(genes_with_CpGs)],
  CpG_start = start(CpG_grange)[queryHits(genes_with_CpGs)],
  CpG_end = end(CpG_grange)[queryHits(genes_with_CpGs)],
  transcript_chr = seqnames(transcripts)[subjectHits(genes_with_CpGs)],
  transcript_start = start(transcripts)[subjectHits(genes_with_CpGs)],
  transcript_end = end(transcripts)[subjectHits(genes_with_CpGs)],
  transcript_id = transcripts$transcript_id[subjectHits(genes_with_CpGs)],
  gene_id = transcripts$gene_id[subjectHits(genes_with_CpGs)]
)

#how many CpGs are represented
length(CpG_transcript_annot$transcript_id)

# how many CpGs are in these genes?
cpg_counts <- CpG_transcript_annot %>%
  group_by(gene_id) %>%
  summarise(num_cpgs = n())

hist <- ggplot(cpg_counts, aes(x = num_cpgs)) +
  geom_histogram(binwidth = 3, fill = "#1f78b4", color = "black") +
  scale_x_continuous(limits = c(0, max(cpg_counts$num_cpgs))) +
  labs(x = "Number of CpGs per gene passing coverage filtering",
       y = "Count of genes") +
  theme_minimal()

save_ggplot(hist, "../output_WGBS/figures/1_cpghist")

# Calculate summary stats
summary_stats <- cpg_counts %>%
  summarise(
    min_cpgs = min(num_cpgs),
    median_cpgs = median(num_cpgs),
    mean_cpgs = mean(num_cpgs),
    max_cpgs = max(num_cpgs),
    total_genes = n(),
    one_cpg = sum(num_cpgs == 1)
  )

# Create annotation text
annotation_text <- paste0(
  "Total genes: ", summary_stats$total_genes, "\n",
  "Min CpGs: ", summary_stats$min_cpgs, "\n",
  "Median: ", summary_stats$median_cpgs, "\n",
  "Mean: ", round(summary_stats$mean_cpgs,2), "\n",
  "Max CpGs: ", summary_stats$max_cpgs, "\n",
  "Genes with 1 CpG: ", summary_stats$one_cpg, "(",(round(summary_stats$one_cpg/summary_stats$total_genes*100)),"%)")

# Plot histogram with annotation box
hist +
  annotate("text", 
           x = Inf, y = Inf, 
           label = annotation_text, 
           hjust = 1.1, vjust = 1.1, 
           size = 4, 
           fontface = "italic",
           color = "black")
```

```{r}
# Find overlaps between methylated loci and transcripts, averaged by gene
CpGs_in_genes <- regionCounts(meth_filter_destrand, regions=transcripts)
CpGs_in_genes$gene_id <- transcripts$gene_id[match(paste(CpGs_in_genes$chr, CpGs_in_genes$start, CpGs_in_genes$end), paste(seqnames(transcripts), start(transcripts), end(transcripts)))]

#how many genes are represented
nrow(CpGs_in_genes)

#does this match the above?
length(unique(CpG_transcript_annot$transcript_id))

percent_meth <- percMethylation(CpGs_in_genes)
percent_meth <- as.data.frame(percent_meth)
percent_meth$gene_id <- CpGs_in_genes$gene_id 
percent_meth <- percent_meth %>% select(gene_id, everything())

percent_meth <- percent_meth %>% rowwise() %>%
  mutate(percent_meth_ALL = mean(c_across(starts_with("LCM")), na.rm = TRUE)) %>%
  ungroup()

percent_meth <- percent_meth %>% rowwise() %>%
  mutate(Oral = mean(c_across(meta$Sample[meta$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  mutate(Aboral = mean(c_across(meta$Sample[meta$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()
  
percent_meth_long <- percent_meth %>% pivot_longer(
                                      cols = c(Oral,Aboral),
                                      names_to = "Tissue",
                                      values_to = "tissue_percent_meth"
)
```

Now, combine the genes that contain methylated loci that passed filtering with the genes that were found to be expressed in our RNA-seq dataset (14464). 

```{r}
# number of genes containing CpGs
nrow(percent_meth)

# number of expressed genes
nrow(DESeq)

expressed_genes_percent_meth <- merge(percent_meth, DESeq, by.x = "gene_id", by.y = "query")

nrow(expressed_genes_percent_meth)

plot_data <- merge(percent_meth_long, DESeq, by.x = "gene_id", by.y = "query")

ggplot(percent_meth, aes(x = percent_meth_ALL)) +
  geom_histogram(bins = 50) +
  theme_minimal() +
  labs(x = "(% gene body methylation", y = "Gene count")
```

### BaseMean: Raw gene expression levels (skewed by highly expressed genes)

```{r}
ggplot(plot_data, aes(y = baseMean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "DeSeq2 BaseMean", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

ggplot(plot_data, aes(y = baseMean, x = tissue_percent_meth,color=Tissue)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "DeSeq2 BaseMean", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()
```

### Log2 of BaseMean: Transformed gene expression levels

```{r}
ggplot(plot_data, aes(y = log2(baseMean), x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Log2(DeSeq2 BaseMean)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

ggplot(plot_data, aes(y = log2(baseMean), x = tissue_percent_meth,color=Tissue)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Log2(DeSeq2 BaseMean)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()
```

### Absolute value of Log2FoldChange: Relationship between overall methylation of a gene and whether or not it is differentially expressed 

```{r}
ggplot(plot_data, aes(y = abs(log2FoldChange), x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +  stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Abs(DeSeq2 Log2FoldChange)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

ggplot(plot_data, aes(y = abs(log2FoldChange), x = tissue_percent_meth,color=Tissue)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +  stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Abs(DeSeq2 Log2FoldChange)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

# Create the plot
ggplot(plot_data, aes(x = percent_meth_ALL, y = log2FoldChange)) +
  geom_point(aes(color = padj < 0.05)) +
  #geom_text_repel(aes(label = ifelse(padj < 0.05, gene_id, "")), max.overlaps = 20) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("grey", "red")) +
  labs(x = "Average CpG % methylation of gene", 
       y = "RNA-seq: Log2FoldChange",
       title = "LFC and DEGs (red) vs. % CpG Methylation",
       color = "Significant DEGs") +
  theme_minimal()
```

### Read counts/vsd transformed counts

RNA metadata:

```{r}
meta_RNA <- read.csv("../data_RNA/LCM_RNA_metadata.csv") %>%
            dplyr::arrange(Sample) %>%
            mutate(across(c(Tissue, Fragment, Section_Date, LCM_Date), factor)) # Set variables as factors 

meta_RNA$Tissue <- factor(meta_RNA$Tissue, levels = c("OralEpi","Aboral")) #we want OralEpi to be the baseline

oral_samples <- meta_RNA$Sample[meta_RNA$Tissue == "OralEpi"]
aboral_samples <- meta_RNA$Sample[meta_RNA$Tissue == "Aboral"]
```

```{r}
vsd <- read.csv("../output_RNA/differential_expression/vsd_expression_matrix.csv", row.names = 1)
vsd$vst_mean <- rowMeans(vsd)
vsd$gene_id <- rownames(vsd)

vsd <- vsd %>%select(gene_id,everything())
vsd <- vsd %>% rowwise() %>%
  mutate(Oral = mean(c_across(all_of(oral_samples)), na.rm = TRUE),
         Aboral = mean(c_across(all_of(aboral_samples)), na.rm = TRUE)) %>% 
 # mutate(Oral = mean(c_across(meta_RNA$Sample[meta_RNA$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  #mutate(Aboral = mean(c_across(meta_RNA$Sample[meta_RNA$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()
  
vsd_long <- vsd %>% pivot_longer(cols = c(Oral,Aboral),
                                      names_to = "Tissue",
                                      values_to = "tissue_vst_mean") %>%
                    mutate(Tissue= factor(Tissue,levels = c("Oral", "Aboral")))

VST_all <- vsd_long %>% ggplot(aes(x = Tissue,y = tissue_vst_mean, fill = Tissue)) + 
                        geom_boxplot() + theme_minimal() + scale_fill_manual(values = c("Aboral" = "mediumpurple1", "Oral" = "palegreen3")) +
                        labs(x = "Tissue", y = "VST Expression", 
                             title = "VST expression of all expressed genes by tissue type",
                             caption = "n = 14,464 genes") 

save_ggplot(VST_all, "../output_WGBS/figures/4_VST_all_boxplot")
```

```{r}
plot_data_tissue <- merge(percent_meth_long, vsd_long, by = c("gene_id", "Tissue")) %>% mutate(Tissue= factor(Tissue,levels = c("Oral", "Aboral")))

vst_percmeth <- ggplot(plot_data_tissue, aes(y = tissue_vst_mean, x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Gene Methylation vs Expression") + scale_color_manual(values = c("Aboral" = "mediumpurple1", "Oral" = "palegreen4")) +
  theme_minimal()

save_ggplot(vst_percmeth, "../output_WGBS/figures/5_VST_tissue_percentmeth")

plot_data <- merge(percent_meth, vsd, by = c("gene_id"))

ggplot(plot_data, aes(y = vst_mean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()
```

```{r}
model <- lm(tissue_vst_mean ~ tissue_percent_meth * Tissue, data = plot_data_tissue)

summary(model)
```

```{r}
# Full model
full_model <- lm(vst_mean ~ percent_meth_ALL, data = plot_data)

# Per-tissue models
oral_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = filter(plot_data_tissue, Tissue == "Oral"))
aboral_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = filter(plot_data_tissue, Tissue == "Aboral"))

summary(full_model)$r.squared
summary(oral_model)$r.squared
summary(aboral_model)$r.squared
```

```{r}
# Model 1: simple model, no tissue info
full_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = plot_data_tissue)

# Model 2: interaction model, includes tissue and interaction term
interaction_model <- lm(tissue_vst_mean ~ tissue_percent_meth * Tissue, data = plot_data_tissue)

summary(full_model)
summary(interaction_model)
anova(full_model, interaction_model)
```

### Above but only for DEGS


RNA metadata:

```{r}
vsd_de <- vsd %>% filter(gene_id %in% DE_05_OralEpi$query)
vsd_de_long <- vsd_de %>% pivot_longer(cols = c(Oral,Aboral),
                                      names_to = "Tissue",
                                      values_to = "tissue_vst_mean")
```

```{r}
plot_data_tissue <- merge(percent_meth_long, vsd_long, by = c("gene_id", "Tissue")) %>% filter(gene_id %in% DE_05$query)

ggplot(plot_data_tissue, aes(y = tissue_vst_mean, x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (ALL): Gene Methylation vs Expression") +
  theme_minimal()

plot_data <- merge(percent_meth, vsd, by = c("gene_id"))

ggplot(plot_data, aes(y = vst_mean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (ALL): Gene Methylation vs Expression") +
  theme_minimal()

plot_data_tissue <- merge(percent_meth_long, vsd_long, by = c("gene_id", "Tissue")) %>% filter(gene_id %in% DE_05_OralEpi$query)

ggplot(plot_data_tissue, aes(y = tissue_vst_mean, x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (Higher in Oral): Gene Methylation vs Expression") +
  theme_minimal()

plot_data <- merge(percent_meth, vsd, by = c("gene_id"))

ggplot(plot_data, aes(y = vst_mean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (Higher in Oral): Gene Methylation vs Expression") +
  theme_minimal()

plot_data_tissue <- merge(percent_meth_long, vsd_long, by = c("gene_id", "Tissue")) %>% filter(gene_id %in% DE_05_Aboral$query)

ggplot(plot_data_tissue, aes(y = tissue_vst_mean, x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (Higher in Aboral): Gene Methylation vs Expression") +
  theme_minimal()

plot_data <- merge(percent_meth, vsd, by = c("gene_id"))

ggplot(plot_data, aes(y = vst_mean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "VST Expression", 
       title = "Differentially Expressed Genes (Higher in Aboral): Gene Methylation vs Expression") +
  theme_minimal()
```


```{r}
model <- lm(tissue_vst_mean ~ tissue_percent_meth * Tissue, data = plot_data_tissue)

summary(model)
```
```{r}
# Full model
full_model <- lm(vst_mean ~ percent_meth_ALL, data = plot_data)

# Per-tissue models
oral_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = filter(plot_data_tissue, Tissue == "Oral"))
aboral_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = filter(plot_data_tissue, Tissue == "Aboral"))

summary(full_model)$r.squared
summary(oral_model)$r.squared
summary(aboral_model)$r.squared
```

```{r}
# Model 1: simple model, no tissue info
full_model <- lm(tissue_vst_mean ~ tissue_percent_meth, data = plot_data_tissue)

# Model 2: interaction model, includes tissue and interaction term
interaction_model <- lm(tissue_vst_mean ~ tissue_percent_meth * Tissue, data = plot_data_tissue)

summary(full_model)
summary(interaction_model)
anova(full_model, interaction_model)
```

### Extracting table of genes

```{r}
DML_info <- DML_transcript_annot %>% dplyr::select(start,end,DML_qvalue,DML_methdiff,gene_id)
DML_info <- DML_info %>% group_by(gene_id) %>% summarise(n_DML = n(),
                                                         mean_methdiff = mean(DML_methdiff),
                                                         median_methdiff = median(DML_methdiff),
                                                         mean_methdiff_qvalue = mean(DML_qvalue),
                                                         min_methdiff_qvalue = min(DML_qvalue),
                                                         max_methdiff_qvalue = max(DML_qvalue))
  
RNA_CpG <- left_join(percent_meth, DML_info, by = c("gene_id")) # take percent methylated CpGs per gene matrix and add signficance info for any differentially methylated loci
RNA_CpG <- inner_join(RNA_CpG, vsd, by = c("gene_id")) # add gene expression info, vst-transformed expression values
RNA_CpG <- inner_join(RNA_CpG, DESeq, by = join_by(gene_id == query)) # add DEseq results

RNA_CpG <- RNA_CpG %>% 
              rename(mean_percent_meth_ALL = percent_meth_ALL,
                     Aboral_mean_perc_meth = Aboral.x,
                     Aboral_mean_vst_expr = Aboral.y,
                     Oral_mean_perc_meth = Oral.x,
                     Oral_mean_vst_expr =  Oral.y,
                     vst_mean_gene_expression = vst_mean,
                     DEseq_baseMean = baseMean,
                     DEseq_log2FoldChange = log2FoldChange,
                     DEseq_lfcSE = lfcSE,
                     DEseq_pvalue = pvalue,
                     DEseq_padj = padj
              )

SwissProt <- read.delim("../references/annotation/protein-GO.tsv")

RNA_CpG_annot <- RNA_CpG%>% left_join(SwissProt, by = join_by(gene_id == query)) 
write.csv(RNA_CpG_annot, "../output_WGBS/MethylKit_ConvFilt90_RNA_Meth.csv")
```


## Methylated CpG locations

```{r}
library(dplyr)

# Define sequencing error rate and alpha
p_error <- 0.01
alpha <- 0.05

# Function to compute adjusted binomial p-values and flag methylated sites per sample
compute_methylation_flags <- function(data, coverage_col, methylated_col, p_error = 0.01, alpha = 0.05) {
  # Extract vectors
  coverage <- data[[coverage_col]]
  methylated <- data[[methylated_col]]
  
  # Handle zero or NA coverage by setting p-values to 1 (not significant)
  pvals <- rep(1, length(coverage))
  
  valid <- !is.na(coverage) & !is.na(methylated) & coverage > 0
  
  # Calculate p-values for valid sites only
  pvals[valid] <- 1 - pbinom(q = methylated[valid] - 1, size = coverage[valid], prob = p_error)
  
  # Adjust p-values for multiple testing with FDR correction
  pvals_adj <- p.adjust(pvals, method = "fdr")
  
  # Return logical vector: TRUE if site methylated (adjusted p < alpha), else FALSE
  return(pvals_adj < alpha)
}

# Assuming you know how many samples you have, for example 10 samples:
num_samples <- 10
methylated_CpGs <- meth_filter_destrand

for (i in 1:num_samples) {
  coverage_col <- paste0("coverage", i)
  methylated_col <- paste0("numCs", i)
  methylation_flag_col <- paste0("methylated_sample", i)

  methylated_CpGs[[methylation_flag_col]] <- compute_methylation_flags(
    methylated_CpGs, 
    coverage_col = coverage_col, 
    methylated_col = methylated_col, 
    p_error = p_error, 
    alpha = alpha
  )
}

```

```{r}
methylated_CpGs <- getData(methylated_CpGs)

oral_samples <- which(sample %in% meta$Sample[meta$Tissue=="OralEpi"])
aboral_samples <- which(sample %in%meta$Sample[meta$Tissue=="Aboral"])
   
oral_samples <- paste0("methylated_sample", oral_samples) 
aboral_samples <- paste0("methylated_sample", aboral_samples)

methylated_CpGs <- methylated_CpGs %>%
  rowwise() %>%
  mutate(
    methylated_Oral = sum(c_across(all_of(oral_samples))) >= 3,
    methylated_Aboral = sum(c_across(all_of(aboral_samples))) >= 3,
    methylated_overall = (methylated_Oral | methylated_Aboral)
  ) %>%
  ungroup()

nrow(methylated_CpGs)
sum(methylated_CpGs$methylated_Oral)
sum(methylated_CpGs$methylated_Aboral)
sum(methylated_CpGs$methylated_overall)
```
```{r}
ML_grange = as(methylated_CpGs,"GRanges")

# identify all the MLs that are in transcripts
genes_with_MLs <- findOverlaps(ML_grange, transcripts,ignore.strand = TRUE)

# Extract matching transcript information
ML_transcript_annot <- data.frame(
  ML_chr = seqnames(ML_grange)[queryHits(genes_with_MLs)],
  ML_start = start(ML_grange)[queryHits(genes_with_MLs)],
  ML_end = end(ML_grange)[queryHits(genes_with_MLs)],
  methylated_Oral = (ML_grange$methylated_Oral)[queryHits(genes_with_MLs)],
  methylated_Aboral = (ML_grange$methylated_Aboral)[queryHits(genes_with_MLs)],
  methylated_overall = (ML_grange$methylated_overall)[queryHits(genes_with_MLs)],
  transcript_chr = seqnames(transcripts)[subjectHits(genes_with_MLs)],
  transcript_start = start(transcripts)[subjectHits(genes_with_MLs)],
  transcript_end = end(transcripts)[subjectHits(genes_with_MLs)],
  transcript_id = transcripts$transcript_id[subjectHits(genes_with_MLs)],
  gene_id = transcripts$gene_id[subjectHits(genes_with_MLs)]
)

nrow(ML_transcript_annot)

## At this point, ML_transcript_annot == CpG_transcript_annot with extra info about methylation or not. but it is the same list of CpGs

methylated_transcript_annot <- ML_transcript_annot %>% filter(methylated_overall==TRUE)

num_methylated_genic <- nrow(methylated_transcript_annot)
num_methylated_genic
num_methylated_total <- sum(ML_grange$methylated_overall)
num_methylated_total
percent_genic <- num_methylated_genic / num_methylated_total * 100
percent_genic

percent_intergenic <- 100 - percent_genic
percent_intergenic

num_methylated_genic_oral <- methylated_transcript_annot %>% filter(methylated_Oral) %>% nrow()
num_methylated_oral <- sum(ML_grange$methylated_Oral)
percent_genic_oral <- num_methylated_genic_oral / num_methylated_oral * 100

num_methylated_genic_aboral <- methylated_transcript_annot %>% filter(methylated_Aboral) %>% nrow()
num_methylated_aboral <- sum(ML_grange$methylated_Aboral)
percent_genic_aboral <- num_methylated_genic_aboral / num_methylated_aboral * 100

percent_genic_oral
percent_genic_aboral
```


```{r}
ML_grange = as(methylated_CpGs,"GRanges")

# identify all the MLs that are in transcripts
exons_with_MLs <- findOverlaps(ML_grange, exons,ignore.strand = TRUE)

# Extract matching transcript information
ML_exon_annot <- data.frame(
  ML_chr = seqnames(ML_grange)[queryHits(exons_with_MLs)],
  ML_start = start(ML_grange)[queryHits(exons_with_MLs)],
  ML_end = end(ML_grange)[queryHits(exons_with_MLs)],
  methylated_Oral = (ML_grange$methylated_Oral)[queryHits(exons_with_MLs)],
  methylated_Aboral = (ML_grange$methylated_Aboral)[queryHits(exons_with_MLs)],
  methylated_overall = (ML_grange$methylated_overall)[queryHits(exons_with_MLs)],
  exon_chr = seqnames(exons)[subjectHits(exons_with_MLs)],
  exon_start = start(exons)[subjectHits(exons_with_MLs)],
  exon_end = end(exons)[subjectHits(exons_with_MLs)],
  transcript_id = exons$transcript_id[subjectHits(exons_with_MLs)],
  gene_id = exons$gene_id[subjectHits(exons_with_MLs)]
)

nrow(ML_exon_annot)
```

```{r}
genic_CpGs <- ML_transcript_annot
exonic_CpGs <- ML_exon_annot 

# Counts:
num_methylated_total <- sum(ML_grange$methylated_overall)
num_methylated_total
num_genic <- sum(genic_CpGs$methylated_overall)
num_genic
num_intergenic <- num_methylated_total - num_genic
num_intergenic
num_exonic <- sum(exonic_CpGs$methylated_overall)
num_exonic
num_intronic <- num_genic-num_exonic
num_intronic
```

```{r}
# Inner ring: Genic vs Intergenic
inner_data <- data.frame(
  group = c("Genic", "Intergenic"),
  count = c(num_genic, num_intergenic)
)

inner_data <- inner_data %>%
  mutate(
    fraction = count / sum(count),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label = paste0(group, "\n", round(fraction * 100, 1), "%")
  )

# Outer ring: Exonic, Intronic, Intergenic
outer_data <- data.frame(
  subgroup = c("Exonic", "Intronic", "Intergenic"),
  group = c("Genic", "Genic", "Intergenic"),
  count = c(num_exonic, num_intronic, num_intergenic)
)

outer_data <- outer_data %>%
  mutate(
    fraction = count / sum(count),
    ymax = cumsum(fraction),
    ymin = c(0, head(ymax, n = -1)),
    label = paste0(subgroup, "\n", count)
  )

region_colors <- c(
  "Intergenic" = "#f7a884",   # orange
  "Exonic"     = "#2268ae",   # dark blue
  "Intronic"   = "#d4e6f5",   # light blue
  "Genic"      = "#808080"    # gray (for inner ring genic portion)
)

venn <- ggplot() +
  # Outer ring (Exonic, Intronic, Intergenic)
  geom_rect(data = outer_data, aes(
    ymin = ymin, ymax = ymax,
    xmin = 1, xmax = 2,
    fill = subgroup
  ), color = "white") +

  # Inner ring (Genic, Intergenic)
  geom_rect(data = inner_data, aes(
    ymin = ymin, ymax = ymax,
    xmin = 0, xmax = 1,
    fill = group
  ), color = "white") +

  coord_polar(theta = "y") +
  xlim(c(0, 2)) +
  theme_void() +
  geom_text(data = outer_data, aes(
    x = 1.5,
    y = (ymin + ymax) / 2,
    label = label
  ), size = 4) +
  geom_text(data = inner_data, aes(
    x = 0.5,
    y = (ymin + ymax) / 2,
    label = label
  ), size = 4, fontface = "bold") +
  scale_fill_manual(values = region_colors) +
  ggtitle("Nested Donut: Methylated CpGs by Genomic Region") +
  theme(legend.position = "none")

save_ggplot(venn, "../output_WGBS/figures/2_cpgvenn")
```


```{r}
all_CpGs <- data.frame(
  region = c("Intergenic", "Exonic", "Intronic"),
  total_CpGs = c(nrow(methylated_CpGs)-nrow(genic_CpGs), nrow(exonic_CpGs), nrow(genic_CpGs)-nrow(exonic_CpGs)),
  methylated_CpGs = c(num_intergenic, num_exonic, num_intronic)
)

# Calculate percentage
all_CpGs <- all_CpGs %>%
  mutate(region = factor(region, levels = c("Intronic","Exonic","Intergenic")),
         percent_methylated = (methylated_CpGs / total_CpGs) * 100)

# Define custom colors
region_colors <- c(
  "Intergenic" = "#f7a884",   # orange
  "Exonic"     = "#2268ae",   # dark blue
  "Intronic"   = "#d4e6f5"    # light blue
)

# Plot
bar <- ggplot(all_CpGs, aes(y = region, x = percent_methylated, fill = region)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = paste0(round(percent_methylated, 1), "%")),
            hjust = -0.1, size = 5) +
  scale_fill_manual(values = region_colors) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(
    title = "Methylated CpGs as % of All CpGs by Region",
    x = "% Methylated",
    y = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")

save_ggplot(bar, "../output_WGBS/figures/3_cpgbar")
```

### All DMLs boxplots

```{r}
percent_meth <- percMethylation(meth_filter_destrand)
percent_meth <- as.data.frame(percent_meth)
percent_meth$chr <- meth_filter_destrand$chr
percent_meth$start <- meth_filter_destrand$start
percent_meth$end <- meth_filter_destrand$end
percent_meth$strand <- meth_filter_destrand$strand
percent_meth <- percent_meth %>% select(c(chr,start,end,strand), everything())


percent_meth_DML <- inner_join(percent_meth,DMLs) %>% 
                                rename("DML_qvalue"=qvalue) %>% 
                                rename("DML_pvalue"=pvalue)%>% 
                                rename("DML_methdiff"=meth.diff)
percent_meth_DML_ingene <- inner_join(percent_meth,DML_transcript_annot) 

percent_meth_DML_tissue <- percent_meth_DML %>% rowwise() %>%
  mutate(Oral = mean(c_across(meta$Sample[meta$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  mutate(Aboral = mean(c_across(meta$Sample[meta$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()

percent_meth_DML_ingene_tissue <- percent_meth_DML_ingene %>% rowwise() %>%
  mutate(Oral = mean(c_across(meta$Sample[meta$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  mutate(Aboral = mean(c_across(meta$Sample[meta$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()
  
percent_meth_DML_long <- percent_meth_DML %>%
  pivot_longer(cols = starts_with("LCM_"), 
               names_to = "Sample", 
               values_to = "Percent_Methylation") %>%
  left_join((meta %>% select(c(Sample,Tissue))), by = "Sample") %>%
  filter(!is.na(Percent_Methylation)) %>% mutate(Tissue= factor(Tissue,levels = c("OralEpi", "Aboral")))

boxplot_percmeth_DMLs <- ggplot(percent_meth_DML_long, aes(x = Tissue, y = Percent_Methylation, fill = Tissue)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  labs(title = "Percent Methylation at DMLs by Tissue",
       x = "Tissue Type", y = "Percent Methylation") +
  theme_minimal() + scale_fill_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3"))+
  theme(legend.position = "none") 

save_ggplot(boxplot_percmeth_DMLs, "../output_WGBS/figures/8_boxplot_percmeth_DMLs")


boxplot_percmeth_DMLs_LOG <- ggplot(percent_meth_DML_long, aes(x = Tissue, y = Percent_Methylation, fill = Tissue)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +scale_y_log10() +
  labs(title = "Percent Methylation at DMLs by Tissue",
       x = "Tissue Type", y = "Log10(Percent Methylation)") +
  theme_minimal() + scale_fill_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3")) +
  theme(legend.position = "none")

save_ggplot(boxplot_percmeth_DMLs_LOG, "../output_WGBS/figures/9_boxplot_percmeth_DMLs_LOG")

percent_meth_DML_ingene_long <- percent_meth_DML_ingene %>%
  pivot_longer(cols = starts_with("LCM_"), 
               names_to = "Sample", 
               values_to = "Percent_Methylation") %>%
  left_join((meta %>% select(c(Sample,Tissue))), by = "Sample") %>%
  filter(!is.na(Percent_Methylation)) %>% mutate(Tissue= factor(Tissue,levels = c("OralEpi", "Aboral")))

boxplot_percmeth_DMLs_ingenes <- ggplot(percent_meth_DML_ingene_long, aes(x = Tissue, y = Percent_Methylation, fill = Tissue)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
  labs(title = "DMLs in genes: Percent Methylation at DMLs by Tissue",
       x = "Tissue Type", y = "Percent Methylation") +
  theme_minimal() + scale_fill_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3")) +
  theme(legend.position = "none")

save_ggplot(boxplot_percmeth_DMLs_ingenes, "../output_WGBS/figures/10_boxplot_percmeth_DMLs_ingenes")


boxplot_percmeth_DMLs_ingenes_LOG <- ggplot(percent_meth_DML_ingene_long, aes(x = Tissue, y = Percent_Methylation, fill = Tissue)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 1) +scale_y_log10() +
  labs(title = "DMLs in genes: Percent Methylation at DMLs by Tissue",
       x = "Tissue Type", y = "Log10(Percent Methylation)") +
  theme_minimal() + scale_fill_manual(values = c("Aboral" = "mediumpurple1", "OralEpi" = "palegreen3")) +
  theme(legend.position = "none")

save_ggplot(boxplot_percmeth_DMLs_ingenes_LOG, "../output_WGBS/figures/11_boxplot_percmeth_DMLs_ingenes_LOG")
```
```{r}
library(patchwork)

# Combine non-log plots
combined_boxplot_percmeth_DMLs <- boxplot_percmeth_DMLs + boxplot_percmeth_DMLs_ingenes +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "none")  # ensures consistent theme

# Save non-log version
ggsave("../output_WGBS/figures/12_combined_boxplot_percmeth_DMLs.png",
       combined_boxplot_percmeth_DMLs, width = 10, height = 5, dpi = 300)

# Combine log-transformed plots
combined_boxplot_percmeth_DMLs_LOG <- boxplot_percmeth_DMLs_LOG + boxplot_percmeth_DMLs_ingenes_LOG +
  plot_layout(ncol = 2, guides = "collect") &
  theme(legend.position = "none")

# Save log version
ggsave("../output_WGBS/figures/13_combined_boxplot_percmeth_DMLs_LOG.png",
       combined_boxplot_percmeth_DMLs_LOG, width = 10, height = 5, dpi = 300)

```


