---
title: "09-MethylKit, filtered for >90% conversion efficiency"
author: "Zoe Dellaert"
date: "2025-02-09"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
always_allow_html: true
---

## MethylKit - Reads filtered for >90% conversion efficiency

I am identifying differentially methylated loci using methylkit based on [Yaamini Venkataraman's code](https://osf.io/u46xj)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("methylKit")
require("tidyverse")
require("vegan")
require("gplots")
require("ggplot2")
require("ggrepel")
require("ggpmisc")
require("dichromat")
require("readr")
require("genomationData")
require("genomation")

sessionInfo() #provides list of loaded packages and version of R.
```


```{r}
meta <- read.csv("../data_WGBS/LCM_WGBS_metadata.csv", sep = ",", header = TRUE) %>%
  mutate(Section_Date = as.character(Section_Date), LCM_Date = as.character(LCM_Date),DNA_Extraction_Date = as.character(DNA_Extraction_Date))

meta <- meta %>% arrange(Sample)
```

```{r}
#file_list <- list.files("/scratch3/workspace/zdellaert_uri_edu-shared/methylseq_V3_bwa_test/bwameth/deduplicated/min_efficiency_test",pattern = "^min_90.*_CpG.methylKit$",  full.names = TRUE, include.dirs = FALSE)

file_list <- list.files("../output_WGBS/methylseq_V3_bwa_test/methyldackel/min_efficiency_test_new",pattern = "^min_90.*_CpG.methylKit$",  full.names = TRUE, include.dirs = FALSE)

sample <- gsub("_CpG.methylKit", "", basename(file_list) )
sample <- gsub("min_90_", "", sample)

sample == meta$Sample #the files and metadata are in the same order

tissue <- meta$Tissue
tissue_binary <- gsub("Aboral", "1", tissue)
tissue_binary <- gsub("OralEpi", "0", tissue_binary)
tissue_binary <- as.numeric(tissue_binary)
fragment <- meta$Fragment

methylObj=methRead(as.list(file_list),
           sample.id = as.list(sample),
           assembly = "Pacuta",
           treatment = tissue_binary,
           context = "CpG",
           mincov = 5
           )

# save(methylObj, file = "../output_WGBS/MethylKit.RData")
```


```{r}
#load("../output_WGBS/MethylKit.RData")

getMethylationStats(methylObj[[2]],plot=FALSE,both.strands=FALSE)
getMethylationStats(methylObj[[2]],plot=TRUE,both.strands=FALSE)
getCoverageStats(methylObj[[2]],plot=TRUE,both.strands=FALSE)
getCoverageStats(methylObj[[2]],plot=TRUE,both.strands=TRUE)
```

```{r}
filtered_methylObj=filterByCoverage(methylObj,lo.count=5,lo.perc=NULL,
                                      hi.count=NULL,hi.perc=99.9)

filtered_methylObj_norm <- filtered_methylObj %>% methylKit::normalizeCoverage(.)
```


```{r}
meth_filter=methylKit::unite(filtered_methylObj_norm)
meth_filter_destrand=methylKit::unite(filtered_methylObj_norm,destrand = TRUE)
 
clusterSamples(meth_filter, dist="correlation", method="ward", plot=TRUE)
clusterSamples(meth_filter_destrand, dist="correlation", method="ward", plot=TRUE)

PCASamples(meth_filter)
PCASamples(meth_filter_destrand)
```

```{r}
getCorrelation(meth_filter_destrand,plot=TRUE)
```

### batch effects 

```{r}
as=assocComp(mBase=meth_filter_destrand,dplyr::select(meta,c("PCR_ReAmp_Cycles", "Fragment")))
as
```


### other possible filtering

```{r}
# get percent methylation matrix
pm=percMethylation(meth_filter_destrand)

# calculate standard deviation of CpGs
sds=matrixStats::rowSds(pm)

# Visualize the distribution of the per-CpG standard deviation
# to determine a suitable cutoff
hist(sds, breaks = 100)

# keep only CpG with standard deviations larger than 2%
meth <- meth_filter_destrand[sds > 2]

# This leaves us with this number of CpG sites
nrow(meth_filter_destrand)
nrow(meth)
```


### Identify DML

```{r}
DMLStats_Tissue <- methylKit::calculateDiffMeth(meth_filter_destrand, overdispersion = "MN", test = "Chisq", mc.cores = 8) #Calculate differential methylation statistics and include covariate information.

head(DMLStats_Tissue) #Look at differential methylation output
```

```{r}
# Filter DMRs with q-value < 0.05
significant_dmg <- getData(DMLStats_Tissue[DMLStats_Tissue$qvalue < 0.1, ])

# Create a data frame for plotting
plot_data <- data.frame(
  chr = significant_dmg$chr,
  start = significant_dmg$start,
  meth.diff = significant_dmg$meth.diff
)

# Count the number of positive and negative methylation differences
positive_count <- sum(significant_dmg$meth.diff > 0)
negative_count <- sum(significant_dmg$meth.diff < 0)

# Plot with counts added to the quadrants
ggplot(plot_data, aes(x = start, y = meth.diff)) +
  geom_point(alpha = 0.5) +  # Set alpha to reduce point transparency
  theme_minimal() +
  labs(title = "Significant Differentially Methylated Regions (q-value < 0.05)",
       x = "Genomic Position (start)",
       y = "Methylation Difference (%)") +
  theme(legend.position = "none") +  # Remove the legend
  # Add the count of positive and negative methylation differences as text annotations
  annotate("text", x = Inf, y = Inf, label = paste("Positive:", positive_count), 
           hjust = 1.1, vjust = 1.1, size = 4, color = "blue") +
  annotate("text", x = Inf, y = -Inf, label = paste("Negative:", negative_count), 
           hjust = 1.1, vjust = -0.1, size = 4, color = "red")


dml_df <- as.data.frame(DMLStats_Tissue)

# Volcano plot
ggplot(dml_df, aes(x = meth.diff, y = -log10(qvalue))) +
  geom_point(alpha = 0.5) +
  geom_vline(xintercept = c(-5, 5), linetype = "dashed", color = "red") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") +
  labs(title = "Differentially Methylated Loci (DMLs)",
       x = "Methylation Difference (%)",
       y = "-log10(q-value)") +
  theme_minimal()
```


```{r}
DMLs <- methylKit::getMethylDiff(DMLStats_Tissue, difference = 2, qvalue = 0.05) #Identify DML based on difference threshold

length(DMLs$chr) #DML

head(DMLs)
```
## Further look at genome wide methylation

```{r}
diffMethPerChr(DMLStats_Tissue, meth.cutoff = 2, qvalue.cutoff = 0.05,cex.names=.75)
```


### Annotation

```{r}
require("genomationData")
require("genomation")

gff.file = "../references/Pocillopora_acuta_HIv2.gtf"
gff = gffToGRanges(gff.file)
head(gff)
```


```{r}
exons = gffToGRanges(gff.file, filter = "exon")

transcripts = gffToGRanges(gff.file, filter = "transcript")
```

```{r}
DML_grange = as(DMLs,"GRanges")
head(DML_grange)
```


```{r}
transcripts = gffToGRanges(gff.file, filter = "transcript")

# Find overlaps between DMLs and transcripts
overlaps_transcripts <- findOverlaps(DML_grange, transcripts,ignore.strand = TRUE)

# Extract matching transcript information
DML_transcript_annot <- data.frame(
  DML_chr = seqnames(DML_grange)[queryHits(overlaps_transcripts)],
  DML_start = start(DML_grange)[queryHits(overlaps_transcripts)],
  DML_end = end(DML_grange)[queryHits(overlaps_transcripts)],
  DML_qvalue = (DML_grange$qvalue)[queryHits(overlaps_transcripts)],
  DML_methdiff = (DML_grange$meth.diff)[queryHits(overlaps_transcripts)],
  transcript_chr = seqnames(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_start = start(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_end = end(transcripts)[subjectHits(overlaps_transcripts)],
  transcript_id = transcripts$transcript_id[subjectHits(overlaps_transcripts)],
  gene_id = transcripts$gene_id[subjectHits(overlaps_transcripts)]
)

# View first few rows
head(DML_transcript_annot)
```

## Are any DMGs DMLs?

```{r}
#load in DESeq results
DESeq <- read.csv("../output_RNA/differential_expression/DESeq_results.csv", header = TRUE) %>% dplyr::rename("query" ="X")

#make dataframes of just differentially expressed genes for each LFC direction
DE_05_Aboral <- DESeq %>% filter(padj < 0.05 & log2FoldChange > 0)
DE_05_OralEpi <- DESeq %>% filter(padj < 0.05& log2FoldChange < 0)
DE_05 <- DESeq %>% filter(padj < 0.05)

DML_transcript_annot[DML_transcript_annot$transcript_id %in% DE_05$query,]

DE_05[DE_05$query %in% DML_transcript_annot$transcript_id,]
```


```{r}
plot_data <- merge(DML_transcript_annot, DESeq, by.x = "transcript_id", by.y = "query")

ggplot(plot_data, aes(x = DML_methdiff, y = log2FoldChange)) +
  geom_point(aes(color = padj < 0.05), size = 3) +
  geom_text_repel(aes(label = ifelse(padj < 0.05, transcript_id, "")), max.overlaps = 20) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("grey", "red")) +
  labs(x = "Methylation Difference (%)", 
       y = "RNA-seq: Log2FoldChange",
       title = "Differentially methylated loci (DMLs) vs. Gene Expression",
       color = "Significant DEGs") +
  theme_minimal()
```
```{r}
# Find overlaps between methylated loci and transcripts
MLs_in_genes <- regionCounts(meth_filter_destrand, regions=transcripts)

#MLs_in_genes <- getData(MLs_in_genes)
MLs_in_genes$gene_id <- transcripts$gene_id[match(paste(MLs_in_genes$chr, MLs_in_genes$start, MLs_in_genes$end), paste(seqnames(transcripts), start(transcripts), end(transcripts)))]

percent_meth <- percMethylation(MLs_in_genes)
percent_meth <- as.data.frame(percent_meth)
percent_meth$gene_id <- MLs_in_genes$gene_id 

percent_meth <- percent_meth %>% rowwise() %>%
  mutate(percent_meth_ALL = mean(c_across(starts_with("LCM")), na.rm = TRUE)) %>%
  ungroup()

percent_meth <- percent_meth %>% rowwise() %>%
  mutate(Oral = mean(c_across(meta$Sample[meta$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  mutate(Aboral = mean(c_across(meta$Sample[meta$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()
  
percent_meth_long <- percent_meth %>% pivot_longer(
                                      cols = c(Oral,Aboral),
                                      names_to = "Tissue",
                                      values_to = "tissue_percent_meth"
)
```

```{r}
plot_data <- merge(percent_meth, DESeq, by.x = "gene_id", by.y = "query")

ggplot(plot_data, aes(y = baseMean, x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "DeSeq2 BaseMean", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

ggplot(plot_data, aes(y = log2(baseMean), x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Log2(DeSeq2 BaseMean)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

ggplot(plot_data, aes(y = abs(log2FoldChange), x = percent_meth_ALL)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm") +  stat_poly_eq(use_label("eq", "R2")) +
  labs(x = "Average CpG % methylation of gene", y = "Abs(DeSeq2 Log2FoldChange)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()

# Create the plot
ggplot(plot_data, aes(x = percent_meth_ALL, y = log2FoldChange)) +
  geom_point(aes(color = padj < 0.05)) +
  #geom_text_repel(aes(label = ifelse(padj < 0.05, gene_id, "")), max.overlaps = 20) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("grey", "red")) +
  labs(x = "Average CpG % methylation of gene", 
       y = "RNA-seq: Log2FoldChange",
       title = "Differentially methylated loci (DMLs) vs. Gene Expression",
       color = "Significant DEGs") +
  theme_minimal()
```

```{r}
plot_data <- merge(percent_meth_long, DESeq, by.x = "gene_id", by.y = "query")

ggplot(plot_data, aes(y = abs(log2FoldChange), x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "Abs(DeSeq2 Log2FoldChange)", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()
```

### Read counts

RNA metadata:

```{r}
meta_RNA <- read.csv("../data_RNA/LCM_RNA_metadata.csv") %>%
            dplyr::arrange(Sample) %>%
            mutate(across(c(Tissue, Fragment, Section_Date, LCM_Date), factor)) # Set variables as factors 

meta_RNA$Tissue <- factor(meta_RNA$Tissue, levels = c("OralEpi","Aboral")) #we want OralEpi to be the baseline
```


```{r}
filtered_counts_RNA <- read.csv("../output_RNA/differential_expression/filtered_counts.csv") 
rownames(filtered_counts_RNA) <- filtered_counts_RNA$X
filtered_counts_RNA <- filtered_counts_RNA %>% dplyr::select(-X)
filtered_counts_RNA$sum <- rowSums(filtered_counts_RNA)
filtered_counts_RNA$gene_id <- rownames(filtered_counts_RNA)

filtered_counts_RNA <- filtered_counts_RNA %>% rowwise() %>%
  mutate(Oral = mean(c_across(meta_RNA$Sample[meta_RNA$Tissue=="OralEpi"]), na.rm = TRUE)) %>%
  mutate(Aboral = mean(c_across(meta_RNA$Sample[meta_RNA$Tissue=="Aboral"]), na.rm = TRUE)) %>%
  ungroup()
  
filtered_counts_RNA_long <- filtered_counts_RNA %>% pivot_longer(
                                      cols = c(Oral,Aboral),
                                      names_to = "Tissue",
                                      values_to = "tissue_mean_counts"
)
```

```{r}
plot_data <- merge(percent_meth_long, filtered_counts_RNA_long, by = c("gene_id", "Tissue"))

ggplot(plot_data, aes(y = log2(tissue_mean_counts), x = tissue_percent_meth, color=Tissue)) +
  geom_point(alpha = 0.5) + geom_smooth(method = "lm") + stat_poly_eq(use_label("eq", "R2"))+
  labs(x = "Average CpG % methylation of gene", y = "Mean counts, all samples", 
       title = "Gene Methylation vs Expression") +
  theme_minimal()
```