---
title: "10-CpG-Methylation-fb"
author: "Zoe Dellaert"
date: "2025-02-17"
output:
  github_document:
    toc: true
    toc_depth: 3
    number_sections: true
    html_preview: true
always_allow_html: true
---

## CpG Methylation analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.width = 8,  
  fig.height = 6, 
  dpi = 300   
)
```

## Managing Packages Using Renv

To run this code in my project using the renv environment, run the following lines of code

```{r, eval=FALSE}
install.packages("renv") #install the package on the new computer (may not be necessary if renv bootstraps itself as expected)
renv::restore() #reinstall all the package versions in the renv lockfile
```

## Load packages

```{r packages}
require("tidyverse")
require("ggplot2")
require("gtools")

sessionInfo() #provides list of loaded packages and version of R.
```


```{r}
meta <- read.csv("../data_WGBS/LCM_WGBS_metadata.csv", sep = ",", header = TRUE) %>%
  mutate(Section_Date = as.character(Section_Date), LCM_Date = as.character(LCM_Date),DNA_Extraction_Date = as.character(DNA_Extraction_Date))

meta_simple <- meta %>% select(Sample, Fragment, Tissue, PCR_ReAmp_Cycles)
```


"Bisulfite conversion efficiency was also estimated from coral alignments as the ratio of the sum of unmethylated cytosines in CHG and CHH context to the sum of methylated and unmethyl- ated cytosines in CHG and CHH."


## All cytosines - CpG, CHG, and CHH intersection with gene bodies

```{r}
#get list of all sample output directories and extract sample names
sample_directories <- list.files("~/zdellaert_uri_edu-shared/methylseq_bwa_fb/methyldackel", pattern = "_quant", full.names = TRUE, include.dirs = TRUE)
samples <- gsub("_quant","",basename(sample_directories))

# make an empty data frame to store methylation data and conversion efficiency data
all_methylation_data <- data.frame()
conversion_eff_data <- data.frame()

for (sample in samples) {
  output_dir <- paste0("~/zdellaert_uri_edu-shared/methylseq_bwa_fb/methyldackel/", sample, "_quant")

  # Read percent methylation data for CpG, CHG, and CHH contexts
  CpG <- read.table(file.path(output_dir, "gene_body_CpG_methylation.txt"), header=FALSE)
  CHG <- read.table(file.path(output_dir, "gene_body_CHG_methylation.txt"), header=FALSE)
  CHH <- read.table(file.path(output_dir, "gene_body_CHH_methylation.txt"), header=FALSE)

  # label dataframe with context
  CpG$context <- "CpG"
  CHG$context <- "CHG"
  CHH$context <- "CHH"
  
  # combine into one
  methylation_data <- rbind(CpG, CHG, CHH)
  colnames(methylation_data) <- c("scaffold", "transcript_start", "transcript_end", "transcript_id", "methylation", "context")
  methylation_data$methylation <- as.numeric(methylation_data$methylation)
  methylation_data$sample <- sample

  # Append methylation data for this sample
  all_methylation_data <- rbind(all_methylation_data, methylation_data)

  # read in file length info for CHH and CHG total counts and unmethyated counts
  count_table <- read.table(file.path(output_dir, "efficiency.txt"), header=FALSE)
  counts <- as.numeric(count_table$V1)

  unmethylated_CHG <- counts[1]
  unmethylated_CHH <- counts[2]
  total_CHG <- counts[3]
  total_CHH <- counts[4]

  # Sum of unmethylated cytosines
  unmethylated_total <- unmethylated_CHG + unmethylated_CHH

  # Sum of total cytosines (methylated + unmethylated)
  total_cytosines <- total_CHG + total_CHH

  # Bisulfite conversion efficiency calculation
  conversion_efficiency <- unmethylated_total / total_cytosines
  print(paste(sample, "Bisulfite Conversion Efficiency: ", conversion_efficiency))

  conversion_eff_data <- rbind(conversion_eff_data, data.frame(sample=sample, efficiency=conversion_efficiency))
}

write.csv(all_methylation_data, "../output_WGBS/methylseq_bwa_fb/gene_body_methylation.csv",row.names = FALSE)
write.csv(conversion_eff_data, "../output_WGBS/methylseq_bwa_fb/conversion_efficiency.csv",row.names = FALSE)
```

Analyze and plot:

```{r}
conversion_eff_data <- read.csv( "../output_WGBS/methylseq_bwa_fb/conversion_efficiency.csv", sep = ",", header = TRUE) 
all_methylation_data <- read.csv( "../output_WGBS/methylseq_bwa_fb/gene_body_methylation.csv", sep = ",", header = TRUE) 

conversion_eff_data <- conversion_eff_data %>%
  left_join(meta_simple,by = join_by(sample == Sample)) %>%
  mutate(sample = fct_relevel(sample, conversion_eff_data$sample[mixedorder(conversion_eff_data$sample)])) 

all_methylation_data <- all_methylation_data %>% left_join(meta_simple,by = join_by(sample == Sample)) %>%
  mutate(sample = fct_relevel(sample, (unique(all_methylation_data$sample)[mixedorder(unique(all_methylation_data$sample))]))) 

# Plot methylation distributions by context and sample
ggplot(all_methylation_data, aes(x=sample, y=methylation, fill=Fragment)) +
  geom_boxplot() +
  facet_grid(~context, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="Gene Body Methylation at CpG, CHG, and CHH loci", y="Mean Methylation (%)")

ggplot(conversion_eff_data, aes(x=sample, y=efficiency, fill=Tissue)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

ggplot(conversion_eff_data, aes(x=sample, y=efficiency, fill=PCR_ReAmp_Cycles)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

summary <- all_methylation_data %>%
  group_by(sample, context) %>%
  summarize(mean_methylation = mean(methylation, na.rm = TRUE)) %>% filter(context =="CpG") %>% left_join(conversion_eff_data)

ggplot(summary, aes(x=mean_methylation, y=efficiency, color=Fragment, label=sample)) +
  geom_point() +
    geom_text(nudge_x = -2.5) + 
  #geom_smooth() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y="bisulfite conversion efficiency")
```

##  5X Coverage Cytosines ONLY

```{r}
#get list of all sample output directories and extract sample names
sample_directories <- list.files("~/zdellaert_uri_edu-shared/methylseq_bwa_fb/methyldackel", pattern = "_5Xquant", full.names = TRUE, include.dirs = TRUE)
samples <- gsub("_5Xquant","",basename(sample_directories))

# make an empty data frame to store methylation data and conversion efficiency data
all_methylation_data_5x <- data.frame()
conversion_eff_data_5x <- data.frame()

for (sample in samples) {
  output_dir <- paste0("~/zdellaert_uri_edu-shared/methylseq_bwa_fb/methyldackel/", sample, "_5Xquant")
  
  # Read percent methylation data for CpG, CHG, and CHH contexts
  CpG <- read.table(file.path(output_dir, "gene_body_CpG_methylation.txt"), header=FALSE)
  CHG <- read.table(file.path(output_dir, "gene_body_CHG_methylation.txt"), header=FALSE)
  CHH <- read.table(file.path(output_dir, "gene_body_CHH_methylation.txt"), header=FALSE)

  # label dataframe with context
  CpG$context <- "CpG"
  CHG$context <- "CHG"
  CHH$context <- "CHH"
  
  # combine into one
  methylation_data <- rbind(CpG, CHG, CHH)
  colnames(methylation_data) <- c("scaffold", "transcript_start", "transcript_end", "transcript_id", "methylation", "context")
  methylation_data$methylation <- as.numeric(methylation_data$methylation)
  methylation_data$sample <- sample

  # Append methylation data for this sample
  all_methylation_data_5x <- rbind(all_methylation_data_5x, methylation_data)

  # read in file length info for CHH and CHG total counts and unmethyated counts
  count_table <- read.table(file.path(output_dir, "efficiency.txt"), header=FALSE)
  counts <- as.numeric(count_table$V1)

  unmethylated_CHG <- counts[1]
  unmethylated_CHH <- counts[2]
  total_CHG <- counts[3]
  total_CHH <- counts[4]

  # Sum of unmethylated cytosines
  unmethylated_total <- unmethylated_CHG + unmethylated_CHH

  # Sum of total cytosines (methylated + unmethylated)
  total_cytosines <- total_CHG + total_CHH

  # Bisulfite conversion efficiency calculation
  conversion_efficiency <- unmethylated_total / total_cytosines
  print(paste(sample, "Bisulfite Conversion Efficiency: ", conversion_efficiency))

  conversion_eff_data_5x <- rbind(conversion_eff_data_5x, data.frame(sample=sample, efficiency=conversion_efficiency))
}

write.csv(all_methylation_data_5x, "../output_WGBS/methylseq_bwa_fb/gene_body_methylation_5x.csv",row.names = FALSE)
write.csv(conversion_eff_data_5x, "../output_WGBS/methylseq_bwa_fb/conversion_efficiency_5x.csv",row.names = FALSE)
```

Analyze and plot:

```{r}
conversion_eff_data_5x <- read.csv( "../output_WGBS/methylseq_bwa_fb/conversion_efficiency_5x.csv", sep = ",", header = TRUE) 
all_methylation_data_5x <- read.csv( "../output_WGBS/methylseq_bwa_fb/gene_body_methylation_5x.csv", sep = ",", header = TRUE) 

conversion_eff_data_5x <- conversion_eff_data_5x %>%
  left_join(meta_simple,by = join_by(sample == Sample)) %>%
  mutate(sample = fct_relevel(sample, conversion_eff_data_5x$sample[mixedorder(conversion_eff_data_5x$sample)])) 

all_methylation_data_5x <- all_methylation_data_5x %>% left_join(meta_simple,by = join_by(sample == Sample)) %>%
  mutate(sample = fct_relevel(sample, (unique(all_methylation_data_5x$sample)[mixedorder(unique(all_methylation_data_5x$sample))]))) 

# Plot methylation distributions by context and sample
ggplot(all_methylation_data_5x, aes(x=sample, y=methylation, fill=Fragment)) +
  geom_boxplot() +
  facet_grid(~context, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="5X coverage-loci ONLY - Gene Body Methylation at CpG, CHG, and CHH loci", y="Mean Methylation (%)")

ggplot(conversion_eff_data_5x, aes(x=sample, y=efficiency, fill=Tissue)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="5X coverage-loci ONLY - Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

ggplot(conversion_eff_data_5x, aes(x=sample, y=efficiency, fill=PCR_ReAmp_Cycles)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="5X coverage-loci ONLY - Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

summary <- all_methylation_data_5x %>%
  group_by(sample, context) %>%
  summarize(mean_methylation = mean(methylation, na.rm = TRUE)) %>% filter(context =="CpG") %>% left_join(conversion_eff_data_5x)

ggplot(summary, aes(x=mean_methylation, y=efficiency, color=Fragment, label=sample)) +
  geom_point() +
  geom_text(nudge_x = -2.5) + 
  #geom_smooth() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   labs(title="5X coverage-loci ONLY" , y="bisulfite conversion efficiency")

```


## plots side by side 

```{r}
ggplot(all_methylation_data, aes(x=sample, y=methylation, fill=Fragment)) +
  geom_boxplot() +
  facet_grid(~context, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="Gene Body Methylation at CpG, CHG, and CHH loci", y="Mean Methylation (%)")

ggplot(all_methylation_data_5x, aes(x=sample, y=methylation, fill=Fragment)) +
  geom_boxplot() +
  facet_grid(~context, scales = "free") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title="5X coverage-loci ONLY - Gene Body Methylation at CpG, CHG, and CHH loci", y="Mean Methylation (%)")

ggplot(conversion_eff_data, aes(x=sample, y=efficiency, fill=Tissue)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

ggplot(conversion_eff_data_5x, aes(x=sample, y=efficiency, fill=Tissue)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="5X coverage-loci ONLY - Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

ggplot(conversion_eff_data, aes(x=sample, y=efficiency, fill=PCR_ReAmp_Cycles)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

ggplot(conversion_eff_data_5x, aes(x=sample, y=efficiency, fill=PCR_ReAmp_Cycles)) +
  geom_bar(stat="identity", color="black") +
  theme_minimal() +
  facet_grid(~Fragment, scales = "free") +
  labs(title="5X coverage-loci ONLY - Bisulfite Conversion Efficiency per Sample", y="Conversion Efficiency", x="Sample") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(0, 1) 

summary <- all_methylation_data %>%
  group_by(sample, context) %>%
  summarize(mean_methylation = mean(methylation, na.rm = TRUE)) %>% filter(context =="CpG") %>% left_join(conversion_eff_data)

ggplot(summary, aes(x=mean_methylation, y=efficiency, color=Fragment, label=sample)) +
  geom_point() +
    geom_text(nudge_x = -2.5) + 
  #geom_smooth() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(y="bisulfite conversion efficiency")

summary <- all_methylation_data_5x %>%
  group_by(sample, context) %>%
  summarize(mean_methylation = mean(methylation, na.rm = TRUE)) %>% filter(context =="CpG") %>% left_join(conversion_eff_data_5x)

ggplot(summary, aes(x=mean_methylation, y=efficiency, color=Fragment, label=sample)) +
  geom_point() +
  geom_text(nudge_x = -2.5) + 
  #geom_smooth() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title="5X coverage-loci ONLY" , y="bisulfite conversion efficiency")

```

